---
title: "Untitled"
output: html_document
date: "2025-09-03"
---
```{r}
source("~/script/lyb_function.R")
library(xvhelper)
library(data.table)
getwd()
```

## 01. olink dataframe combind
```{r}
instance_0_df <- fread("olink_instance_0.NPX.csv")
#instance_2_df <- fread("olink_instance_2.NPX.csv")
#instance_3_df <- fread("olink_instance_3.NPX.csv")
```

```{r}
instance_0_ldf <- instance_0_df |> mutate(ins_index = 0) |> pivot_longer(cols = -c(eid, ins_index), names_to = "protein_id", values_to = "result") %>% na.omit()
instance_2_ldf <- instance_2_df |> mutate(ins_index = 2) |> pivot_longer(cols = -c(eid, ins_index), names_to = "protein_id", values_to = "result") %>% na.omit()
instance_3_ldf <- instance_3_df |> mutate(ins_index = 3) |> pivot_longer(cols = -c(eid, ins_index), names_to = "protein_id", values_to = "result") %>% na.omit()
```


```{r}
olink_df <- instance_0_ldf %>% bind_rows(instance_2_ldf, instance_3_ldf)
```

```{r}
coding143 <- fread("olink_resources/coding143.tsv") %>% mutate(meaning = str_to_lower(str_replace(meaning, ";.*", "")), meaning = str_replace_all(meaning, "-", "_"))
```

```{r}
field_30900_df <- fread("olink_resources/field_30900.csv") %>%
    select(-participant.p30900_i1) %>%
    pivot_longer(cols = -c(participant.eid), names_to = "instance", values_to = "N_proteins") %>%
    filter(!is.na(N_proteins)) %>%
    mutate(instance = str_remove(instance, "participant.p30900_i"))
field_30900_df$instance <- as.double(field_30900_df$instance)

field_30901_df <- fread("olink_resources/field_30901.csv") %>%
    select(-participant.p30901_i1) %>%
    pivot_longer(cols = -c(participant.eid), names_to = "instance", values_to = "PlateID") %>%
    filter(!is.na(PlateID)) %>%
    mutate(instance = str_remove(instance, "participant.p30901_i"),
          PlateID = as.double(PlateID))
field_30901_df$instance <- as.double(field_30901_df$instance)

field_30902_df <- fread("olink_resources/field_30902.csv") %>%
    select(-participant.p30902_i1) %>%
    pivot_longer(cols = -c(participant.eid), names_to = "instance", values_to = "WellID") %>%
    filter(WellID != "") %>%
    mutate(instance = str_remove(instance, "participant.p30902_i"))
field_30902_df$instance <- as.double(field_30902_df$instance)
```

```{r}
field_30900_df |> group_by(instance) |> summarise(n(),mean(N_proteins))
```
```{r}
olink_assay <- fread("olink_resources/olink_assay.dat") %>% mutate(Assay = tolower(Assay))
olink_assay_version <- fread("olink_resources/olink_assay_version.dat") %>% mutate(Assay = tolower(Assay))
olink_batch_number <- fread("olink_resources/olink_batch_number.dat") %>% mutate(PlateID = as.double(PlateID))
olink_limit_of_detection <- fread("olink_resources/olink_limit_of_detection.dat") %>% mutate(Assay = tolower(Assay), PlateID = as.double(PlateID))
olink_panel_lot_number <- fread("olink_resources/olink_panel_lot_number.dat")
olink_processing_start_date <- fread("olink_resources/olink_processing_start_date.dat")  %>% mutate(PlateID = as.double(PlateID))
```

```{r}
olink_full_dataset_df <- olink_df |> 
    left_join(field_30900_df, by = c("eid" = "participant.eid", "ins_index" = "instance")) %>%
    left_join(field_30901_df, by = c("eid" = "participant.eid", "ins_index" = "instance")) %>%
    left_join(field_30902_df, by = c("eid" = "participant.eid", "ins_index" = "instance")) %>%
    left_join(coding143, by = c("protein_id" = "meaning")) %>%
    left_join(olink_limit_of_detection, by = c("protein_id" = "Assay", "ins_index" = "Instance", "PlateID" = "PlateID")) %>%
    left_join(olink_assay, by = c("protein_id" = "Assay")) %>%
    left_join(olink_processing_start_date, by = c("PlateID" = "PlateID", "Panel" = "Panel")) %>%
    left_join(olink_batch_number, by = c("PlateID" = "PlateID")) %>%
    left_join(olink_panel_lot_number, by = c("Batch" = "Batch", "Panel" = "Panel")) %>%
    left_join(olink_assay_version, by = c("Panel_Lot_Nr" = "Panel_Lot_Nr", "protein_id" = "Assay"))
```

```{r}
#saveRDS(olink_full_dataset_df, file = "/lustre/home/acct-medty/medty-c/project/14.CHIP_UKB/06.ukb_olink/ukb_olink.full_dataset.rds")
olink_full_dataset_df <- readRDS("/lustre/home/acct-medty/medty-c/project/14.CHIP_UKB/06.ukb_olink/ukb_olink.full_dataset.rds")
```

## 02. UKB phenotype
```{r}
codings_df <- read_csv("app176137_20251017152832.dataset.codings.csv", show_col_types = FALSE)
data_dict_df <- read_csv("app176137_20251017152832.dataset.data_dictionary.csv", show_col_types = FALSE)
entity_dict_df <- read_csv("app176137_20251017152832.dataset.entity_dictionary.csv", show_col_types = FALSE)

coded_col_df <- xvhelper::merge_coding_data_dict(coding_dict = codings_df, data_dict = data_dict_df)
```

### basic clinical info 
```{r}
ukb_data_df <- read_csv("ukb_cohort_data.csv", show_col_types = FALSE)
ukb_data_df.date<- read_csv("ukb_cohort_data2.csv", show_col_types = FALSE)
ukb_data_df.pc <- read_csv("GWAS_data/ukb.gwas_pheno.csv", show_col_types = FALSE)
ukb_data_df.pc <- ukb_data_df.pc[,grepl("eid|p22009",colnames(ukb_data_df.pc))]
ukb_data_df <- ukb_data_df |> inner_join(ukb_data_df.date, by = "participant.eid") |> select(-"participant.p23104_i0")
ukb_data_df <- ukb_data_df |> inner_join(ukb_data_df.pc, by = "participant.eid")
```

###  medication
```{r}
ukb_data_df.med <- read_csv("Medication/ukb.medication.csv", show_col_types = FALSE)
statin_codes <- c(
  1140861958, # simvastatin
  1141146234, # atorvastatin
  1141192410, # rosuvastatin
  1140888648, # pravastatin
  1140888594  # fluvastatin
)
med_cols <- grep("p20003", colnames(ukb_data_df.med), value = TRUE)
ukb_data_df.med$statin <- apply(ukb_data_df.med[, med_cols], 1, function(x) {
  any(x %in% statin_codes, na.rm = TRUE)
})

ukb_data_df.med$statin <- as.integer(ukb_data_df.med$statin)

antidiabetic_codes <- c(
  1140874646,1140874674,1140874706,1140874744,
  1141152590,
  1140884600,
  1140883066,
  1141168660,1141173882,
  1141171646,1141177600,
  1141189090
)

ukb_data_df.med$antidiabetic <- apply(ukb_data_df.med[, med_cols], 1, function(x) {
  any(x %in% antidiabetic_codes, na.rm = TRUE)
})

ukb_data_df.med$antidiabetic <- as.integer(ukb_data_df.med$antidiabetic)


antiplatelet_codes <- c(
  1140861778, # dipyridamole
  1140861806, # aspirin 75mg tablet
  1140864860, # nu-seals aspirin 75mg e/c tablet
  1140868226, # aspirin
  1141167844, # dipyridamole+aspirin
  1141168318  # clopidogrel
)

anticoagulant_codes <- c(
  1140861588, # enoxaparin
  1140861702, # phenindione
  1140864956, # subcutaneous heparin
  1140881842, # heparin
  1140888204, # dalteparin
  1140888206, # tinzaparin
  1140888266, # warfarin
  1140909770, # acenocoumarol
  1140910832  # sodium warfarin
)

antithrombotic_code <- c(antiplatelet_codes, anticoagulant_codes)
ukb_data_df.med$antithrombotic <- apply(ukb_data_df.med[, med_cols], 1, function(x) {
  any(x %in% antithrombotic_code, na.rm = TRUE)
})
ukb_data_df.med$antithrombotic <- as.integer(ukb_data_df.med$antithrombotic)

sum(ukb_data_df.med$statin)
sum(ukb_data_df.med$antidiabetic)
sum(ukb_data_df.med$antithrombotic)
```


```{r}
ukb_med_codes <- read.delim("Medication/coding4.tsv")
med.df <- data.frame(coding = c(statin_codes,antidiabetic_codes,antithrombotic_code),
           type = c( rep("statin",length(statin_codes)),
                     rep("antidiabetic",length(antidiabetic_codes)),
                     rep("antithrombotic",length(antithrombotic_code)) )
)

med.df$meaning <- ukb_med_codes$meaning[match(med.df$coding,ukb_med_codes$coding)]


med_long <- ukb_data_df.med |>
  select(participant.eid, all_of(med_cols)) |>
  pivot_longer(
    cols = all_of(med_cols),
    names_to = "med_field",
    values_to = "med_code"
  ) |>
  filter(!is.na(med_code), med_code != "") |>
  mutate(med_code = as.numeric(med_code)) |>
  distinct(participant.eid, med_code, .keep_all = TRUE)

med_long <- med_long |>  filter(participant.eid %in% chip_mat.phen.df$eid)
med_count <- med_long |> count(med_code)

med.df$count <- med_count$n[match(med.df$coding,med_count$med_code)]
colnames(med.df) <-c("UKB medication code","Medication category", "Medication name", "Number of users")
med.df

write.table(med.df, "chip_suppl_table_med.tsv", quote = FALSE, sep = "\t", row.names = FALSE)
```

```{r}
ukb_data_df <- ukb_data_df |>
    xvhelper::decode_single(coded_col_df) |>
    xvhelper::decode_multi_purrr(coded_col_df) |>
    xvhelper::decode_column_names(coded_col_df, r_clean_names = FALSE)

ukb_data_df$`Participant ID` <- as.integer(ukb_data_df$`Participant ID`)
```

```{r}
ukb_data_df |> group_by(`Past tobacco smoking | Instance 0`,`Current tobacco smoking | Instance 0`,`Ever smoked | Instance 0` ) |>  summarise(n())
#eid BMI Age Sex Smoking
```
```{r}
smoking_eid <- (ukb_data_df |> filter(`Past tobacco smoking | Instance 0` == "Smoked on most or all days" | `Current tobacco smoking | Instance 0` == "Yes, on most or all days"))$`Participant ID`
```

### sub
```{r}
ukb_data_df.s <-  ukb_data_df[,c("Participant ID","Sex","Age when attended assessment centre | Instance 0","Body mass index (BMI) | Instance 0","Date of attending assessment centre | Instance 0","Ever smoked | Instance 0")]
colnames(ukb_data_df.s) <- c("eid","Sex","Age","BMI","Date","Smoking")
ukb_data_df.pc <- ukb_data_df.pc |> rename_with(~ gsub("participant.","", .x)) |>  rename_with(~ gsub("p22009_a","pc", .x)) 
ukb_data_df.s <- ukb_data_df.s |>  inner_join(ukb_data_df.pc, by = c("eid"))
ukb_data_df.s <- ukb_data_df.s |>  inner_join(ukb_data_df.med[,c("participant.eid","statin","antidiabetic","antithrombotic")], by = c("eid" = "participant.eid" ))

#ukb_data_df.s$Smoking <- "no"
#ukb_data_df.s[ukb_data_df.s$eid %in% smoking_eid,]$Smoking <- "yes"

ukb_data_df.s |> group_by(Sex,Smoking) |> summarise(n())
ukb_data_df.s |> group_by(statin,antidiabetic) |> summarise(n())
```
```{r}
ukb_data_df  |> count(Sex)
ggplot(ukb_data_df, aes(x=`Age at recruitment`)) + geom_density(fill = "#08537C", alpha = 0.2, color = "#08537C") + theme_test()
```
### Blood count
```{r}
ukb_blood_count_df <- read_csv("ukb.blood_count.csv", show_col_types = FALSE)
data_dict_df.blood_count <- data_dict_df |> mutate(ent_field = paste0(entity,".",name)) |> filter(ent_field %in% colnames(ukb_blood_count_df)) |> mutate(ent_name = paste0(title,"(",units,")"))

data_dict_df.blood_count$ent_name_s  <- c("eid", "WBC (×10⁹/L)", "RBC (×10¹²/L)", "Hb (g/dL)", "HCT (%)", "MCV (fL)", "MCH (pg)", "MCHC (g/dL)", "RDW (%)", "PLT (×10⁹/L)", "PCT (%)", "MPV (fL)", "PDW (%)", "Lymphocytes (×10⁹/L)", "Monocytes (×10⁹/L)", "Neutrophils (×10⁹/L)", "Eosinophils (×10⁹/L)", "Basophils (×10⁹/L)", "NRBC (×10⁹/L)", "Lymphocytes (%)", "Monocytes (%)", "Neutrophils (%)", "Eosinophils (%)", "Basophils (%)", "NRBC (%)", "Reticulocytes (%)", "Reticulocytes (×10¹²/L)", "Mean Reticulocyte Volume (fL)", "Mean Sphered Cell Volume (fL)", "Immature Reticulocyte Fraction (ratio)", "High Light Scatter Reticulocytes (%)", "High Light Scatter Reticulocytes (×10¹²/L)")

data_dict_df.blood_count$ent_id  <- c("eid", "WBC_10e9_L", "RBC_10e12_L", "Hb_g_dL", "HCT_pct", "MCV_fL", "MCH_pg", "MCHC_g_dL", "RDW_pct", "PLT_10e9_L", "PCT_pct", "MPV_fL","PDW_pct","Lymphocytes_10e9_L","Monocytes_10e9_L","Neutrophils_10e9_L","Eosinophils_10e9_L","Basophils_10e9_L","NRBC_10e9_L","Lymphocytes_pct","Monocytes_pct","Neutrophils_pct","Eosinophils_pct","Basophils_pct","NRBC_pct","Reticulocytes_pct","Reticulocytes_10e12_L","Mean_Reticulocyte_Volume_fL","Mean_Sphered_Cell_Volume_fL","Immature_Reticulocyte_Fraction_ratio","High_Light_Scatter_Reticulocytes_pct","High_Light_Scatter_Reticulocytes_10e12_L")

index <- match(colnames(ukb_blood_count_df),data_dict_df.blood_count$ent_field)
colnames(ukb_blood_count_df) <- data_dict_df.blood_count$ent_id[index]
ukb_blood_count_df <- ukb_blood_count_df[,data_dict_df.blood_count$ent_id]
colnames(ukb_blood_count_df) 
```

```{r}
ukb_blood_count_df <- ukb_blood_count_df[rowSums(is.na(ukb_blood_count_df)) < 31,]
```

### Serum biochemistry
```{r}
tmp.df.list <- lapply(0:9, function(x){
    df <- read.csv(paste0("Bio_chemistry/ukb.bio_chemistry.0",x,".csv"),na.strings = c("", "NA"))
    return(df)
})

ukb.biochem.df <- purrr::reduce(tmp.df.list, full_join, by = "participant.eid")
rm(tmp.df.list)

data_dict_df.biochem <- data_dict_df |> mutate(ent_field = paste0(entity,".",name)) |> 
  filter(ent_field %in% colnames(ukb.biochem.df)) |> filter(type %in% c("string","float")) |> 
  filter(name != "p30897_i0") |>  mutate(ent_name = paste0(title,"(",units,")"))

data_dict_df.biochem$ent_id <- c("eid", "ALB","ALP","ALT","APOA","APOB","AST","DBIL", "UREA","CALC","CHOL","CREA","CRP","CYSC","GGT","GLU","HbA1c","HDL","IGF-1","LDL","LP-a","OEST","PHOS","RF","SHBG","TBIL","TEST","TP","TRIG","UA","VITD")

data_dict_df.biochem$ent_name[1] <- "eid"
data_dict_df.biochem$ent_name <- sub(" \\| Instance 0","",data_dict_df.biochem$ent_name)
data_dict_df.biochem$ent_name <- gsub(" ","_",data_dict_df.biochem$ent_name)

index <- match(data_dict_df.biochem$ent_field,colnames(ukb.biochem.df))
ukb.biochem.df <- ukb.biochem.df[,index]
#colnames(ukb.biochem.df) <- data_dict_df.biochem$ent_id
colnames(ukb.biochem.df) <- data_dict_df.biochem$ent_name
```

```{r}
ukb.biochem.df <- ukb.biochem.df[rowSums(is.na(ukb.biochem.df)) < 30,]
```

### NMR metabolomics
```{r}
ukb_metabolic_df <- read_csv("ukb.NMR_meta.csv", show_col_types = FALSE)

data_dict_df.metabolic <- data_dict_df |> mutate(ent_field = paste0(entity,".",name)) |> filter(ent_field %in% colnames(ukb_metabolic_df)) |> mutate(ent_name = paste0(title,"(",units,")"))

data_dict_df.metabolic |> group_by(units) |> summarise(n())

ent_name.list <- gsub("\\| Instance 0","",data_dict_df.metabolic$ent_name)
ent_name.list <- gsub(" percentage","",ent_name.list)
ent_name.list <- gsub("percent","%",ent_name.list)
ent_name.list <- gsub(" \\(ratio\\)","",ent_name.list)
ent_name.list <- gsub(" \\(degree\\)","",ent_name.list)
ent_name.list[1] <- "eid"
data_dict_df.metabolic$ent_name <- ent_name.list

index <- match(colnames(ukb_metabolic_df),data_dict_df.metabolic$ent_field)
colnames(ukb_metabolic_df) <- data_dict_df.metabolic$ent_name[index]
ukb_metabolic_df <- ukb_metabolic_df[,data_dict_df.metabolic$ent_name]

#colnames(ukb_metabolic_df)
table(rowSums(is.na(ukb_metabolic_df)))
ukb_metabolic_df <- ukb_metabolic_df[rowSums(is.na(ukb_metabolic_df)) < 251 * (1 - 0.8),]
dim(ukb_metabolic_df)
```


### ICD10

```{r}
#data_ICD10_A <- ukb.icd10.df |>
#  filter(grepl("^A", `Diagnoses - main ICD10`)) |>
#  count(`Diagnoses - main ICD10`) |>
#  arrange(desc(n)) |>
#  gt::gt()

##########input
#ukb.icd10.date.df <- read.delim2("Disease_record/ukb.main_ICD10.csv", sep = ",")
ukb.icd10.date.df <- read.delim2("Disease_record/ukb.ICD10.csv", sep = ",")
ukb.icd10.date.df <- ukb.icd10.date.df[rowSums(ukb.icd10.date.df == "") < 260,]


##########separate icd10
ukb.icd10.df  <- ukb_data_df |>
  dplyr::select(`Participant ID`, `Diagnoses - ICD10`) |>
  tidyr::separate_rows(`Diagnoses - ICD10`, sep="\\|") |> 
  filter(`Diagnoses - ICD10` != "NA") 
  
ukb.icd10.df <- ukb.icd10.df |> group_by(`Participant ID`) |> mutate(array = row_number() -1 ) |> ungroup()

######### process date
ukb.icd10.date.df <- ukb.icd10.date.df[,-2] |>  pivot_longer(cols = starts_with("participant.p"), names_to = "array", values_to = "date") |> filter(date != "") 
#ukb.icd10.date.df$array <- gsub("participant.p41262_a","",ukb.icd10.date.df$array)
ukb.icd10.date.df$array <- gsub("participant.p41280_a","",ukb.icd10.date.df$array)
ukb.icd10.date.df$array <- as.integer(ukb.icd10.date.df$array)

setdiff(paste(ukb.icd10.date.df$participant.eid, ukb.icd10.date.df$array),paste(ukb.icd10.df$`Participant ID`,ukb.icd10.df$array))


####### merge date
ukb.icd10.df <- ukb.icd10.df |> inner_join(ukb.icd10.date.df, by = c("Participant ID" = "participant.eid", "array" = "array"))

####### separate ICD10 & Diagnoses
ukb.icd10.df <- ukb.icd10.df |> separate(`Diagnoses - ICD10`, into = c("ICD10","Diagnosis"), sep = " ", extra = "merge", fill = "right")
colnames(ukb.icd10.df) <- c("eid","ICD10","Diagnosis","array","date")

####### add assessment date
ukb.icd10.df <- ukb.icd10.df |> inner_join(ukb_data_df[,c("Participant ID","Date of attending assessment centre | Instance 0")], by = c("eid" = "Participant ID"))
ukb.icd10.df$date <- as.Date(ukb.icd10.df$date)
colnames(ukb.icd10.df) <- c("eid","ICD10","Diagnosis","array","date","assess_date")
ukb.icd10.df$interval_days <- ukb.icd10.df$date - ukb.icd10.df$assess_date
```

#### thromb-related
```{r}
  #"Hypercoagulability",            "Thrombophilia/hypercoagulable states",   "^D68\\.(5|6)",
  #"Hypercoagulability",            "Essential thrombocythaemia/MPN",         "^(D47\\.3|D75\\.2)$",
thromb_groups <- tibble::tribble(
  ~group,                          ~subgroup,                                ~regex,
  "VTE",                           "Pulmonary embolism",                     "^I26",
  "VTE",                           "DVT/phlebitis",                          "^I80",
  "VTE",                           "Other venous thrombosis",                "^(I81|I82)",
  "VTE (pregnancy)",               "Pregnancy/puerperium VTE",               "^O(22\\.3|22\\.5|87\\.1|87\\.3|88\\.2)$",
  "ATE",  "Myocardial infarction/acute coronary thrombosis", "^I2(1|2)|^I24\\.0",
  "ATE",  "Cerebral infarction",                    "^I63",
  "ATE",  "Arterial embolism/thrombosis",           "^I74",
  "ATE",  "Visceral/mesenteric infarction",         "^(K55\\.0|K76\\.3|N28\\.0|D73\\.5)$",
  "VTE",    "Cerebral venous thrombosis",         "^(I67\\.6|G08)$",

  "Other embolism (consider exclude/sensitivity)",
                                  "Fat/air/amniotic embolism",               "^(T79\\.(0|1)|O88\\.(0|1))$"
)

classify_thromb <- function(icd_vec) {
  icd_vec <- toupper(trimws(icd_vec))
  grp  <- rep(NA_character_, length(icd_vec))
  subg <- rep(NA_character_, length(icd_vec))
  for (i in seq_len(nrow(thromb_groups))) {
    hit <- str_detect(icd_vec, thromb_groups$regex[i])
    grp[ hit]  <- thromb_groups$group[i]
    subg[hit]  <- thromb_groups$subgroup[i]
  }
  tibble::tibble(thromb_group = grp, thromb_subgroup = subg)
}


ukb.tromb.stats <- ukb.icd10.df |> filter(grepl("thromb|embol|infarc",Diagnosis, ignore.case = TRUE)) |> count(ICD10,Diagnosis) |> arrange(desc(n))
ukb.tromb.stats <- cbind(ukb.tromb.stats,classify_thromb(ukb.tromb.stats$ICD10))
#write.table(ukb.tromb.stats, "~/ukb.tromb.stats.txt",sep = "\t",row.names = FALSE)

ukb.icd10.thromb.df <- ukb.icd10.df |> filter(grepl("thromb|embol|infarc",Diagnosis, ignore.case = TRUE))
ukb.icd10.thromb.df <- cbind(ukb.icd10.thromb.df,classify_thromb(ukb.icd10.thromb.df$ICD10))
ukb.icd10.thromb.df <- ukb.icd10.thromb.df |> filter(!is.na(thromb_group))
```

```{r}
ukb.icd10.thromb.df |> group_by(ICD10, Diagnosis) |> count() |> arrange(desc(n))
ukb.icd10.thromb.df |> count(thromb_group) |> arrange(desc(n))
ukb.icd10.thromb.df |> count(thromb_subgroup) |> arrange(desc(n))
#ukb.icd10.thromb.df <- ukb.icd10.thromb.df |> filter(thromb_group %in% c("ATE","VTE","Hypercoagulability"))
ukb.icd10.thromb.df <- ukb.icd10.thromb.df |> filter(thromb_group %in% c("ATE","VTE"))
```


#### add Category
```{r}
ukb.icd10.df <- ukb.icd10.df |>
  mutate(
    Category = case_when(
      grepl("^[AB]", ICD10) ~ "Infectious diseases",
      grepl("^C|^D0|^D1|^D2|^D3|^D4", ICD10) ~ "Neoplasms (Tumors)",
      grepl("^D5|^D6|^D7|^D8", ICD10) ~ "Hematologic/Immune",
      grepl("^E", ICD10) ~ "Endocrine, nutrition, metabolism",
      grepl("^F", ICD10) ~ "Mental/Behavioral",
      grepl("^G", ICD10) ~ "Nervous system",
      grepl("^H0|^H1|^H2|^H3|^H4|^H5", ICD10) ~ "Eye diseases",
      grepl("^H6|^H7|^H8|^H9", ICD10) ~ "Ear diseases",
      grepl("^I", ICD10) ~ "Circulatory system",
      grepl("^J", ICD10) ~ "Respiratory system",
      grepl("^K", ICD10) ~ "Digestive system",
      grepl("^L", ICD10) ~ "Skin/subcutaneous",
      grepl("^M", ICD10) ~ "Musculoskeletal/connective",
      grepl("^N", ICD10) ~ "Genitourinary system",
      grepl("^O", ICD10) ~ "Pregnancy/childbirth",
      grepl("^P", ICD10) ~ "Perinatal conditions",
      grepl("^Q", ICD10) ~ "Congenital anomalies",
      grepl("^R", ICD10) ~ "Symptoms/clinical findings",
      grepl("^S|^T", ICD10) ~ "Injury/poisoning",
      grepl("^V|^W|^X|^Y", ICD10) ~ "External causes",
      grepl("^Z", ICD10) ~ "Health services/contacts",
      grepl("^U", ICD10) ~ "Special codes (e.g., COVID-19)",
      TRUE ~ "Unclassified"
    )
  )
```


```{r}
ukb.icd10.df |> count(Category)

ukb.icd10.df |> filter(grepl("S01",ICD10))
```

### Cancer Register
```{r}
ukb.cancer_register.df <- read.csv("Disease_record/ukb.cancer_register.csv", na.strings = c("", "NA"))
problems(ukb.cancer_register.df)
ukb.cancer_register.df <- ukb.cancer_register.df[rowSums(is.na(ukb.cancer_register.df)) < 170,]
```


```{r}
ukb.cancer_register.df <- ukb.cancer_register.df |> 
  xvhelper::decode_single(coded_col_df) |>
  xvhelper::decode_column_names(coded_col_df, r_clean_names = FALSE)

ukb.cancer_register.long.df <- ukb.cancer_register.df |> pivot_longer(cols = -`Participant ID`, names_to = "Field", values_to = "Value", values_transform = list(Value = as.character)) |> filter(!is.na(Value)) |> separate(Field, into = c("Field","Instance"), sep = "\\|", extra = "merge", fill = "right")

ukb.cancer_register.long.df <- ukb.cancer_register.long.df |> pivot_wider(id_cols = c(`Participant ID`, Instance), names_from = Field, values_from = Value)
```


### Death Register
```{r}
ukb.death_register.df <- read.csv("Disease_record/ukb.death_register.csv", na.strings = c("", "NA"))
ukb.death_register.df <- ukb.death_register.df[rowSums(is.na(ukb.death_register.df)) < 41,]
```

```{r}
ukb.death_register.df <- ukb.death_register.df |> 
  xvhelper::decode_single(coded_col_df) |>
  xvhelper::decode_column_names(coded_col_df, r_clean_names = FALSE)

ukb.death_register.long.df <- ukb.death_register.df |> pivot_longer(cols = -`Participant ID`, names_to = "Field", values_to = "Value", values_transform = list(Value = as.character)) |> filter(!is.na(Value)) |> separate(Field, into = c("Field","Instance"), sep = "\\|", extra = "merge", fill = "right")

max((ukb.death_register.long.df |> filter(Field == "Date of death "))$Value)
```

#### Combind
```{r}
ukb.icd10.df |> filter(grepl("C92",ICD10))
ukb.icd10.df |> filter(eid == "1100317")
ukb.cancer_register.long.df |> filter(`Participant ID` == "1100317")
ukb_data_df.s |> filter(eid == "1100317")
```

```{r}
#Age_Death Date_Death
death_df <- ukb.death_register.df[,c("Participant ID","Date of death | Instance 0","Age at death | Instance 0")]
colnames(death_df) <- c("eid","Date_Death", "Age_Death")
death_df$Date_Death <- as.Date(death_df$Date_Death)

ukb_data_df.s$status <- as.numeric(ukb_data_df.s$eid %in% death_df$eid)
ukb_data_df.s <- ukb_data_df.s |> left_join(death_df, by = c("eid"))
```

## 03. UKB CHIP maf 

```{r}
ukb.maf.1.df <- read.table("/lustre/home/acct-medty/medty-c/project/14.CHIP_UKB/05.ukb_call_mut/process_maf.26319/merge/UKB.CH_mut.all.maf.txt", quote = "", sep = "\t", header = TRUE)
ukb.maf.1.df <- ukb.maf.1.df |> filter(WhiteList >1)
ukb.maf.1.df <- ukb.maf.1.df |> filter(Tumor_Sample_Barcode %in% ukb_data_df.s$eid)
ukb.maf.2.df <- read.table("/lustre/home/acct-medty/medty-c/project/14.CHIP_UKB/05.ukb_call_mut/process_maf.26319/merge/Blood2023_Vlasschaert_UKB450679.chip.hg38_multianno.maf", quote = "", sep = "\t", header = TRUE)
ukb.maf.2.df <- ukb.maf.2.df |> mutate(VAF = as.numeric(str_extract(INFO, "(?<=VAF=)[0-9.]+")))


all(ukb.maf.2.df$Tumor_Sample_Barcode %in% ukb_data_df.s$eid)

ukb.maf.1.s.df <- ukb.maf.1.df[,c(1,5:13,15,17:21)];ukb.maf.1.s.df$VAF = ukb.maf.1.df$Mean_VAF; 
#### calc depth
mean_comma_num_int <- function(x) {
  if (is.na(x) || x == "") return(NA_integer_)
  xx <- as.numeric(strsplit(x, ",")[[1]])
  as.integer(round(mean(xx, na.rm = TRUE)))
}
ukb.maf.1.s.df$Depth = sapply(ukb.maf.1.df$DEPTH, mean_comma_num_int) 

ukb.maf.2.s.df <- ukb.maf.2.df[,c(1,5:13,15,17:21)];ukb.maf.2.s.df$VAF =  ukb.maf.2.df$VAF;

dim(ukb.maf.1.s.df)
dim(ukb.maf.2.s.df)

ukb.maf.df <- ukb.maf.1.s.df |> full_join(ukb.maf.2.s.df, by=colnames(ukb.maf.2.s.df)[1:16])
ukb.maf.df <- ukb.maf.df |> mutate(VAF = if_else(is.na(VAF.x),VAF.y,VAF.x))


# UKB 和 in-house 之间关于indel的突变，位置标注差一个bp，导致同义突变重复注释
ukb_repeat.maf <- ukb.maf.df |> filter(is.na(VAF.x) | is.na(VAF.y)) |> count(Hugo_Symbol,AAChange,Tumor_Sample_Barcode,Variant_Classification) |> filter(n >=2, Variant_Classification != "Splice_Site")
dim(ukb_repeat.maf)



# 1. 定义需要去重处理的那部分

get_first_non_na <- function(x){x <- x[!is.na(x)];if (length(x) == 0) return(NA_real_);x[1]}

maf.na <- ukb.maf.df |>
  filter(is.na(VAF.x) | is.na(VAF.y)) |>
  filter(Variant_Classification != "Splice_Site") |>
#  distinct(Hugo_Symbol, AAChange, Tumor_Sample_Barcode, Variant_Classification, .keep_all = TRUE) 
  group_by(Hugo_Symbol, AAChange, Tumor_Sample_Barcode, Variant_Classification) |>
  summarise(across(.cols = -c(VAF.x, VAF.y),.fns = dplyr::first),
            VAF.x = get_first_non_na(VAF.x),VAF.y = get_first_non_na(VAF.y),.groups = "drop")


# 2. 定义其余不需要动的部分
maf.other <- ukb.maf.df |>
  filter(!(is.na(VAF.x) | is.na(VAF.y)) | Variant_Classification == "Splice_Site")


# 3. 合并回来
ukb.maf.df <- bind_rows(maf.other, maf.na)

length(unique(ukb.maf.df$Tumor_Sample_Barcode)) /  469481
dim(ukb.maf.df)
length(unique(ukb.maf.df$Tumor_Sample_Barcode))
```
#### output for suppl tables
```{r}
ukb.maf.sort.df <- ukb.maf.df %>%
  mutate(
    chr_order = str_remove(Chromosome, "^chr"),
    chr_order = case_when(
      chr_order == "X" ~ "23",
      chr_order == "Y" ~ "24",
      chr_order %in% c("M", "MT") ~ "25",
      TRUE ~ chr_order
    ),
    chr_order = as.integer(chr_order)
  ) %>%
  arrange(chr_order, Start_Position) %>%
  select(-chr_order)

ukb.maf.sort.df <- ukb.maf.sort.df |> mutate(CHIP = case_when(
    !is.na(VAF.x) & !is.na(VAF.y) ~ "Shared",
    !is.na(VAF.x) & is.na(VAF.y) ~ "In-house only",
    is.na(VAF.x) & !is.na(VAF.y) ~ "UKB only",
    TRUE ~ NA_character_
  ))

ukb.maf.sort.df |> count(CHIP)

colnames(ukb.maf.1.df)[c(1,5:13,15,17:21)]
colnames(ukb.maf.sort.df)
colnames(ukb.maf.1.df)

ukb.maf.sort.df <- ukb.maf.sort.df |> full_join(ukb.maf.1.df, by=colnames(ukb.maf.1.df)[c(1,5:13,15,17:21)])
write.table(ukb.maf.sort.df,"chip_suppl_table_chip.tsv",quote = FALSE, sep = "\t",row.names = FALSE)
```

### self-defined vs ukb
```{r}
length(unique(ukb.maf.1.s.df$Tumor_Sample_Barcode))
length(unique(ukb.maf.2.s.df$Tumor_Sample_Barcode))
length(unique(c(ukb.maf.1.s.df$Tumor_Sample_Barcode,ukb.maf.2.s.df$Tumor_Sample_Barcode)))
sum(unique(ukb.maf.1.s.df$Tumor_Sample_Barcode) %in% unique(ukb.maf.2.s.df$Tumor_Sample_Barcode))
```
```{r}
chip.defined.list <- list()
chip.defined.list[["union"]] <- unique(unique(c(ukb.maf.1.s.df$Tumor_Sample_Barcode,ukb.maf.2.s.df$Tumor_Sample_Barcode)))
chip.defined.list[["self"]] <- unique(ukb.maf.1.s.df$Tumor_Sample_Barcode)
chip.defined.list[["ukb"]] <- unique(ukb.maf.2.s.df$Tumor_Sample_Barcode)
chip.defined.list[["overlap"]] <- unique(ukb.maf.1.s.df$Tumor_Sample_Barcode)[unique(ukb.maf.1.s.df$Tumor_Sample_Barcode) %in% unique(ukb.maf.2.s.df$Tumor_Sample_Barcode)]
chip.defined.list[["self.only"]] <-unique(ukb.maf.1.s.df$Tumor_Sample_Barcode)[!unique(ukb.maf.1.s.df$Tumor_Sample_Barcode) %in% unique(ukb.maf.2.s.df$Tumor_Sample_Barcode)]
chip.defined.list[["ukb.only"]] <- unique(ukb.maf.2.s.df$Tumor_Sample_Barcode)[!unique(ukb.maf.2.s.df$Tumor_Sample_Barcode) %in% unique(ukb.maf.1.s.df$Tumor_Sample_Barcode)]

sapply(chip.defined.list,length)
```
```{r}
#ukb_wes_all_eid <- readLines("/lustre/home/acct-medty/medty-c/project/14.CHIP_UKB/05.ukb_call_mut/assets/ukb_wes_eid.list")
ukb_wes_all_eid <- eid.ls$WES
chip_defined.mat <- sapply(chip.defined.list, function(eid_set){
  as.integer(ukb_wes_all_eid %in% eid_set)
})
rownames(chip_defined.mat) <- ukb_wes_all_eid
```

#### diff vaf
```{r}
vaf_plot.1 <- data.frame(VAF = (ukb.maf.df |> filter(!is.na(VAF.x)))$VAF,Label = "In-house CHIP")
vaf_plot.2 <- data.frame(VAF = ukb.maf.2.s.df$VAF,Label = "UKB-released CHIP")
#vaf_plot.2 <- data.frame(VAF = (ukb.maf.df |> filter(!is.na(VAF.y)))$VAF,Label = "UKB-released")
vaf_plot.3 <- data.frame(VAF = (ukb.maf.df |> filter(!is.na(VAF.x), is.na(VAF.y)))$VAF, Label = "In-house only CHIP")
vaf_plot.4 <- data.frame(VAF = (ukb.maf.df |> filter(!is.na(VAF.x), !is.na(VAF.y)))$VAF.x, Label = "Shared variants: in-house VAF")
vaf_plot.5 <- data.frame(VAF = (ukb.maf.df |> filter(!is.na(VAF.x), !is.na(VAF.y)))$VAF.y, Label = "Shared variants: UKB-released VAF")
vaf_plot <- rbind(vaf_plot.1,vaf_plot.2)
vaf_plot <- rbind(vaf_plot,vaf_plot.3)
vaf_plot <- rbind(vaf_plot,vaf_plot.4)
vaf_plot <- rbind(vaf_plot,vaf_plot.5)



median_df <- vaf_plot |>
  group_by(Label) |>
  summarise(
    median_vaf = median(VAF, na.rm = TRUE),
    y_pos = quantile(VAF, 0.98, na.rm = TRUE),
    .groups = "drop"
  )

p_f1_a1 <- vaf_plot |>
  ggplot(aes(x = Label, y = VAF, fill = Label)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  geom_text(
    data = median_df,
    aes(
      x = Label,
      y = y_pos,
      label = paste0(round(median_vaf * 100, 1), "%")
    ),
    inherit.aes = FALSE,
    size = 3,
    vjust = -0.3
  ) +
  theme_test() + theme(axis.text.x = element_text(angle = 270,hjust = 1, vjust = 0.5)) + ylim(0,0.5) + 
  scale_fill_manual(values = c("In-house CHIP" = "#0098c9","In-house only CHIP" = "#B5AA0F","Shared variants: in-house VAF" = "#3C5488", "Shared variants: UKB-released VAF" = "#3C5488","UKB-released CHIP" = "#128640"))

p_f1_a1


pdf("chip_f1_a1.pdf",width = 7, height = 6)
p_f1_a1
dev.off()
```
```{r}
shared_df <- ukb.maf.df |> filter(!is.na(VAF.x), !is.na(VAF.y)) 
median_diff <- median(shared_df$VAF.y - shared_df$VAF.x, na.rm = TRUE)
rho <- cor(shared_df$VAF.x, shared_df$VAF.y, method = "pearson", use = "complete.obs")

shared_df <- shared_df |> filter(!abs(VAF.y - VAF.x) > 0.15)

p_f1_a2 <- ggplot(shared_df, aes(x = VAF.x, y = VAF.y)) +
  geom_point_rast(alpha = 0.2, size = 0.6, shape = 16) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.5) +
  geom_smooth( method = "lm",se = TRUE,linewidth = 0.5) +
  annotate("text",x = 0.05,y = 0.45,hjust = 0,
           label = paste0("Pearson's r = ", round(rho, 3),"\nMedian ΔVAF = ", round(median_diff * 100,2), "%")) +
  theme_test() + labs( x = "In-house VAF",y = "UKB-released VAF") +
  ylim(0,0.5) + xlim(0,0.5) 

p_f1_a2

pdf("chip_f1_a2.pdf",width = 5, height = 5)
p_f1_a2
dev.off()
```

#### Depth
```{r}
ukb.maf.1.s.df |> ggplot(aes(Depth)) + geom_density(fill = "#00A087") + theme_test()
#ukb.maf.1.s.df |> mutate(Label = "DEPTH") |> filter(Depth > 20) |> ggplot(aes(Label,Depth)) + geom_violin(fill = "#00A087") + geom_boxplot(width = 0.15)  + scale_y_log10() 

library(dplyr)
library(ggplot2)

# 计算 density 和峰值位置
depth_vec <- ukb.maf.df$Depth
depth_vec <- depth_vec[!is.na(depth_vec)]

den <- density(depth_vec)
peak_df <- data.frame(Depth = den$x[which.max(den$y)], Density = den$y[which.max(den$y)])
peak_df

p_f1_a4 <- ukb.maf.df |> filter(!is.na(Depth)) |>
  ggplot(aes(x = Depth)) +
  geom_density(fill = "#00A087", alpha = 0.6) +
  geom_vline(data = peak_df,  aes(xintercept = Depth), linetype = "dashed") +
  geom_point(data = peak_df,aes(x = Depth, y = Density), shape = 16, size = 3, color = "#A81E2C") +
  annotate("text", x = peak_df$Depth, y = peak_df$Density,
    label = paste0("Peak depth = ", round(peak_df$Depth, 2), "X"),
    vjust = 1, hjust = -0.1, size = 4) +
  theme_test() + labs(x = "Depth",y = "Density")

p_f1_a4


pdf("chip_f1_a4.pdf",width = 4, height = 4)
p_f1_a4
dev.off()
```

#### gene specific vaf

```{r}
ukb.maf.df |> group_by(Hugo_Symbol,AAChange) |> filter(Hugo_Symbol == "NRAS") |> count()
```

```{r}
driver_event_df <- ukb.maf.df |>
  mutate(
    Driver_event = case_when(
      Hugo_Symbol == "DNMT3A" &
        grepl("^p\\.R882", AAChange) ~
        "DNMT3A R882H/C/S",

      Hugo_Symbol == "TET2" &
        Variant_Classification %in% c(
          "Nonsense_Mutation", "Frame_Shift_Del", "Frame_Shift_Ins", "Splice_Site"
        ) ~
        "TET2 truncating/splice",
      
      Hugo_Symbol == "ASXL1" &
        Variant_Classification %in% c(
          "Nonsense_Mutation", "Frame_Shift_Del", "Frame_Shift_Ins", "Splice_Site"
        ) ~
        "ASXL1 truncating/splice",

      Hugo_Symbol == "PPM1D" &
        Variant_Classification %in% c(
          "Nonsense_Mutation", "Frame_Shift_Del", "Frame_Shift_Ins","Splice_Site"
        ) ~
        "PPM1D truncating/splice",

      Hugo_Symbol == "TP53" &
        (
          Variant_Classification %in% c(
            "Nonsense_Mutation", "Frame_Shift_Del", "Frame_Shift_Ins", "Splice_Site"
          ) |
            grepl("^p\\.(R136|R234)", AAChange)
        ) ~
        "TP53 R136/R234/truncating",

      Hugo_Symbol == "JAK2" &
        AAChange == "p.V617F" ~
        "JAK2 V617F",

      Hugo_Symbol == "SRSF2" &
        grepl("^p\\.P95", AAChange) ~
        "SRSF2 P95H/L/R",

      Hugo_Symbol == "SF3B1" &
        AAChange == "p.K700E" ~
        "SF3B1 K700E",

      Hugo_Symbol %in% c("IDH1", "IDH2") &
        (
          (Hugo_Symbol == "IDH1") |
          (Hugo_Symbol == "IDH2")
        ) ~
        "IDH1/2 R132/R140 etc.",

      Hugo_Symbol %in% c("KRAS", "NRAS")  ~
        "KRAS/NRAS G12/G13 etc.",

      TRUE ~ NA_character_
    ),
    Label = case_when(
      !is.na(VAF.x) & !is.na(VAF.y) ~ "Shared",
      !is.na(VAF.x) &  is.na(VAF.y) ~ "In-house only",
       is.na(VAF.x) & !is.na(VAF.y) ~ "UKB-released only",
      TRUE ~ NA_character_
    ),
    Label = factor(
      Label,
      levels = c("In-house only", "Shared", "UKB-released only")
    )
  ) |>
  filter(!is.na(Driver_event), !is.na(Label))

driver_event_count <- driver_event_df |>
  distinct(Tumor_Sample_Barcode, Hugo_Symbol, AAChange, Driver_event, Label, .keep_all = TRUE) |>
  count(Driver_event, Label, name = "n") |>
  group_by(Driver_event) |>
  mutate(total_n = sum(n)) |>
  ungroup() |>
  mutate(
    Driver_event = fct_reorder(Driver_event, total_n)
  )

p_driver_event <- driver_event_count |>
  ggplot(aes(x = Driver_event, y = n, fill = Label)) +
  geom_col(width = 0.75) +
  coord_flip() +
  theme_test() +
  labs(
    x = NULL,
    y = "Number of driver-defining events",
    fill = NULL
  ) + 
  scale_fill_manual(values = c("In-house only" = "#B5AA0F","Shared" = "#3C5488", "UKB-released only" = "#128640"))

p_driver_event

total_label_df <- driver_event_count |>
  group_by(Driver_event) |>
  summarise(total_n = sum(n), .groups = "drop")

p_f1_a3 <- p_driver_event +
  geom_text(
    data = total_label_df,
    aes(x = Driver_event, y = total_n, label = total_n),
    inherit.aes = FALSE,
    hjust = -0.2,
    size = 3
  ) +
  expand_limits(y = max(total_label_df$total_n) * 1.12)
p_f1_a3
pdf("chip_f1_a3.pdf",width = 7, height = 4)
p_f1_a3
dev.off()
```



### maftools
```{r}
library(maftools)
#ukb.maf.df <- read.table("/lustre/home/acct-medty/medty-c/project/14.CHIP_UKB/05.ukb_call_mut/process_maf/UKB.CH_mut.all.sort.maf.txt.gz", quote = "", sep = "\t", header = TRUE)
ukb.maf <- read.maf(ukb.maf.df)
getSampleSummary(ukb.maf)
getGeneSummary(ukb.maf)
```

```{r}
col <- brewer.pal(n=10, name="Paired")
#col <- c("#3C5488","#A81E2C","#B5AA0F","#7BA39D","#B15928","#6A3D9A","#1F78B4","#8491B4","#7E6148","#E64B35","#4DBBD5","#00A087","#A20056","#008B45","#808180","#1B1919")
names(col) = c('Frame_Shift_Del','Missense_Mutation', 'Nonsense_Mutation', 'Multi_Hit', 'Frame_Shift_Ins', 'In_Frame_Ins', 'Splice_Site', 'In_Frame_Del', "Amp", "Del")
plotmafSummary(maf = ukb.maf, rmOutlier = TRUE, addStat = 'median', dashboard = TRUE, titvRaw = FALSE, color = col)

pdf("chip_sf1.pdf", width = 9, height = 5)
plotmafSummary(maf = ukb.maf, rmOutlier = TRUE, addStat = 'median', dashboard = TRUE, titvRaw = FALSE, color = col)
dev.off()
```

```{r}
oncoplot(ukb.maf, top = 20, bgCol = "grey90", borderCol = NULL, color=col,showTumorSampleBarcodes =FALSE, removeNonMutated = FALSE, drawRowBar = TRUE , drawColBar = FALSE, SampleNamefontSize = 0.8, legendFontSize = 1, fontSize = 0.8, titleFontSize = 1.2, cohortSize = 469481)

#pdf("chip_f1b.pdf",width = 12,height = 6,useDingbats = FALSE)
oncoplot(ukb.maf, top = 20, bgCol = "grey90", borderCol = NULL, color=col,showTumorSampleBarcodes =FALSE, removeNonMutated = FALSE, drawRowBar = TRUE , drawColBar = FALSE, SampleNamefontSize = 0.8, legendFontSize = 1,draw_titv = FALSE, fontSize = 0.8, titleFontSize = 1.2, titleText = "CHIP carriers: 31,210 (6.65%); mutations: 34,527")
#dev.off()
```
### vaf
```{r}
chip_genes <- c("DNMT3A","TET2","ASXL1","PPM1D","TP53","SRSF2","JAK2","SF3B1")
ukb.maf.df |> filter(Hugo_Symbol %in% chip_genes) |> group_by(Hugo_Symbol) |> summarise(median(VAF))
ukb.maf.df |> filter(Hugo_Symbol %in% chip_genes) |> 
  mutate(Hugo_Symbol = factor(Hugo_Symbol, levels =  rev(chip_genes))) |> 
  ggplot(aes(y = VAF, x = Hugo_Symbol)) + geom_violin(aes(fill = Hugo_Symbol), show.legend = FALSE) + 
  geom_jitter_rast(width = 0.08, alpha = 0.1) + geom_boxplot(width=0.1,outlier.shape = NA) + 
  theme_test() + ylab("VAF") + xlab("")  + scale_fill_manual(values = rev(mycol[1:8])) + 
  geom_hline(yintercept = c(0.1,0.2),linetype = 2,color = c("#3C5488","#C81E76")) + coord_flip()  + ylim(0,0.5)

pdf("chip_f1c.pdf",width = 4,height = 4,useDingbats = FALSE)
ukb.maf.df |> filter(Hugo_Symbol %in% chip_genes) |> 
  mutate(Hugo_Symbol = factor(Hugo_Symbol, levels =  rev(chip_genes))) |> 
  ggplot(aes(y = VAF, x = Hugo_Symbol)) + geom_violin(aes(fill = Hugo_Symbol), show.legend = FALSE) + 
  geom_jitter_rast(width = 0.08, alpha = 0.1) + geom_boxplot(width=0.1,outlier.shape = NA) + 
  theme_test() + ylab("VAF") + xlab("")  + scale_fill_manual(values = rev(mycol[1:8])) + 
  geom_hline(yintercept = c(0.1,0.2),linetype = 2,color = c("#3C5488","#C81E76")) + coord_flip() + ylim(0,0.5)
dev.off()
```

```{r}
ukb.maf.df |> filter(Hugo_Symbol == "JAK2")  |> count(AAChange)

ukb.maf.df |> filter(Hugo_Symbol %in% chip_genes)|> count(Hugo_Symbol) |>
  mutate(Hugo_Symbol = factor(Hugo_Symbol, levels =  rev(chip_genes))) 
```

### white list
```{r}
wl.maf.df <- read.table("../05.ukb_call_mut/process_maf.26319/WhiteList.CH_mut.merge.maf.txt", quote = "", sep = "\t", header = TRUE)
wl.maf.df <- wl.maf.df |> filter(Hugo_Symbol %in% unique(ukb.maf.df$Hugo_Symbol)) 
```

```{r}
wl.list <- c("Blood2023_Vlasschaert_UKB450679","Nature2020_Bick_TOPMED97691","NG2020_Bolton_MSK24145","NG2022_Kar_UKB200453","NC2023_Brown_UKB198178","NG2024_Bernstein_UKB200618")
#pdf("chip_sf2.pdf",width = 6,height = 6)
lapply(wl.list,function(x){
  center <- x
  #center <- "NG2024_Bernstein_UKB200618"
  tmp.maf.df <- wl.maf.df |> filter(Center  == center)
  n_mut <- format(dim(tmp.maf.df)[1], big.mark = ",", scientific = FALSE)
  tmp.maf <- read.maf(tmp.maf.df)
  oncoplot(tmp.maf, top = 20, bgCol = "grey90", borderCol = NULL, color=col,showTumorSampleBarcodes =FALSE, removeNonMutated = FALSE, drawRowBar = TRUE , drawColBar = FALSE, SampleNamefontSize = 0.8, legendFontSize = 1, fontSize = 0.8, titleFontSize = 1.2, titleText =paste0("CHIP mutation landscape in ", center, " (", n_mut, " mutations)"))
})
#dev.off()
```

```{r}
unique(ukb.maf.df$Hugo_Symbol)

maf_gene.num.df.1 <- wl.maf.df |> filter(Hugo_Symbol %in% unique(ukb.maf.df$Hugo_Symbol)) |> mutate(Genes = if_else(Hugo_Symbol %in% chip_genes, Hugo_Symbol, "Other")) |> group_by(Genes,Center) |> count()
maf_gene.num.df.2 <- ukb.maf.df |> mutate(Genes = if_else(Hugo_Symbol %in% chip_genes, Hugo_Symbol, "Other"), Center = "In_This_Study") |> group_by(Genes,Center) |> count()
maf_gene.num.df.3 <- ukb.maf.1.df |> mutate(Genes = if_else(Hugo_Symbol %in% chip_genes, Hugo_Symbol, "Other"), Center = "self") |> group_by(Genes,Center) |> count()
maf_gene.num.df.4 <- ukb.maf.df |> filter(! is.na(VAF.x) & is.na(VAF.y)) |> mutate(Genes = if_else(Hugo_Symbol %in% chip_genes, Hugo_Symbol, "Other"), Center = "self-only") |> group_by(Genes,Center) |> count()

maf_gene.num.df <- rbind(maf_gene.num.df.1,maf_gene.num.df.2)
maf_gene.num.df <- rbind(maf_gene.num.df,maf_gene.num.df.3)
maf_gene.num.df <- rbind(maf_gene.num.df,maf_gene.num.df.4)
maf_gene.num.df$Genes <- factor(maf_gene.num.df$Genes, levels = c(chip_genes,"Other"))

unique(maf_gene.num.df$Center)

#pdf("chip_f1d.pdf",width = 4,height = 6,useDingbats = FALSE)
maf_gene.num.df |> filter(Center %in% c("In_This_Study","self","self-only","Blood2023_Vlasschaert_UKB450679","Nature2020_Bick_TOPMED97691","NG2020_Bolton_MSK24145","NG2022_Kar_UKB200453","NC2023_Brown_UKB198178","NG2024_Bernstein_UKB200618")) |> mutate(Center2 = factor(Center, levels = c("In_This_Study","self","self-only","Blood2023_Vlasschaert_UKB450679","Nature2020_Bick_TOPMED97691","NG2020_Bolton_MSK24145","NG2022_Kar_UKB200453","NC2023_Brown_UKB198178","NG2024_Bernstein_UKB200618"))) |> ggplot(aes(Center2,n,fill=Genes)) + geom_col(position = "fill") + 
  theme_test() + scale_fill_manual(values = mycol)  + ylab("Ratio") +  xlab("") + 
  theme(axis.text.x = element_text(angle = 90,hjust = 1, vjust = 0.5))
#dev.off()
```

### olink chip maf 
```{r}
ukb_wes_all_eid <- readLines("/lustre/home/acct-medty/medty-c/project/14.CHIP_UKB/05.ukb_call_mut/assets/ukb_wes_eid.list")
ukb_wes_chip_eid <- unique(ukb.maf.df$Tumor_Sample_Barcode)
ukb_wes_all_eid[!(ukb_wes_all_eid %in% ukb_data_df$`Participant ID`)]


ukb_olink_chip_eid <- ukb_wes_chip_eid[ukb_wes_chip_eid %in% instance_0_df$eid]
```
```{r}
olink_chip.maf.df <- ukb.maf.df[match(ukb_olink_chip_eid,ukb.maf.df$Tumor_Sample_Barcode),]
olink_chip.maf <- read.maf(olink_chip.maf.df |> filter(VAF > 0.1))
oncoplot(olink_chip.maf, top = 20, bgCol = "grey90", borderCol = NULL, color=col,showTumorSampleBarcodes =FALSE, removeNonMutated = FALSE, drawRowBar = TRUE , drawColBar = FALSE, SampleNamefontSize = 0.8, legendFontSize = 1, fontSize = 0.8, titleFontSize = 1.2)
#cohortSize = sum(instance_0_df$eid %in% ukb_wes_all_eid)
```

### chip mut dataframe
```{r}
chip_mut_df <- data.frame(eid = ukb.maf.df$Tumor_Sample_Barcode, CHIP = paste0(ukb.maf.df$Hugo_Symbol,":",ukb.maf.df$AAChange), VAF = ukb.maf.df$VAF)
chip_mut_df <- chip_mut_df |> group_by(eid) |> summarise(Mut = paste(CHIP,collapse = ";"),Num = n()) 
chip_mut_df
```

### CHIP clasify by Genes
```{r}
eid.chip.list <- list()
eid.chip.list[["CHIP.all"]] <- unique(ukb.maf.df$Tumor_Sample_Barcode)
eid.chip.list[["CHIP.lar"]] <- unique((ukb.maf.df |> filter(VAF >= 0.1))$Tumor_Sample_Barcode)
eid.chip.list[["CHIP.sma"]] <- eid.chip.list[["CHIP.all"]][ ! eid.chip.list[["CHIP.all"]]  %in% eid.chip.list[["CHIP.lar"]] ]

eid.chip.list[["JAK2.all"]] <- unique((ukb.maf.df |> filter(Hugo_Symbol == "JAK2"))$Tumor_Sample_Barcode)
eid.chip.list[["JAK2.lar"]] <- unique((ukb.maf.df |> filter(VAF >= 0.2) |> filter(Hugo_Symbol == "JAK2"))$Tumor_Sample_Barcode)
eid.chip.list[["JAK2.sma"]] <- eid.chip.list[["JAK2.all"]][ ! eid.chip.list[["JAK2.all"]]  %in% eid.chip.list[["JAK2.lar"]] ]

for (x in c("DNMT3A","TET2","ASXL1","PPM1D","TP53","SRSF2","SF3B1")) {
  all <- unique((ukb.maf.df |> filter(Hugo_Symbol == x))$Tumor_Sample_Barcode)
  large <- unique((ukb.maf.df |> filter(VAF >= 0.1) |> filter(Hugo_Symbol == x))$Tumor_Sample_Barcode)
  small <- all[ ! all %in% large]
  eid.chip.list[[paste0(x,".all")]] <- all
  eid.chip.list[[paste0(x,".lar")]] <- large
  eid.chip.list[[paste0(x,".sma")]] <- small
}

sapply(eid.chip.list,length)
```
```{r}
ukb_wes_all_eid <- readLines("/lustre/home/acct-medty/medty-c/project/14.CHIP_UKB/05.ukb_call_mut/assets/ukb_wes_eid.list")
chip_mat <- sapply(eid.chip.list, function(eid_set){
  as.integer(ukb_wes_all_eid %in% eid_set)
})
rownames(chip_mat) <- ukb_wes_all_eid
```

### CHIP & phenotype
```{r}
chip_mat.phen.df <- as.data.frame(chip_mat)
chip_mat.phen.df$eid <- as.double(rownames(chip_mat.phen.df))
chip_mat.phen.df <- ukb_data_df.s |> inner_join(chip_mat.phen.df, by = c("eid"))
chip_mat.phen.df <- chip_mat.phen.df |> left_join(ukb_blood_count_df, by = c("eid"))
chip_mat.phen.df <- chip_mat.phen.df |> left_join(ukb.biochem.df, by = c("eid"))

#eid.no_icd10 <- chip_mat.phen.df$eid[! chip_mat.phen.df$eid %in% unique(ukb.icd10.df$eid)]
#chip_mat.phen.df <- chip_mat.phen.df |> filter(!eid %in% eid.no_icd10)
```

```{r}
thromb.eid <- unique(ukb.icd10.thromb.df$eid)
chip_mat.phen.df$thromb <- 0
chip_mat.phen.df[chip_mat.phen.df$eid %in% thromb.eid,]$thromb <- 1

#hyter.eid <- (ukb.icd10.thromb.df |> filter(thromb_group == "Hypercoagulability"))$eid
#chip_mat.phen.df$hyter <- 0
#chip_mat.phen.df[chip_mat.phen.df$eid %in% hyter.eid,]$hyter <- 1

ATE.eid <- (ukb.icd10.thromb.df |> filter(thromb_group == "ATE"))$eid
chip_mat.phen.df$ATE <- 0
chip_mat.phen.df[chip_mat.phen.df$eid %in% ATE.eid,]$ATE <- 1

VTE.eid <- (ukb.icd10.thromb.df |> filter(thromb_group == "VTE"))$eid
chip_mat.phen.df$VTE <- 0
chip_mat.phen.df[chip_mat.phen.df$eid %in% VTE.eid,]$VTE <- 1

```


```{r}
chip_mat.phen.df |> group_by(CHIP.all, thromb) |> count()
(3818 / (39787 + 3818)) / (27386 / (398489 + 27386))

chip_mat.phen.df$PLT_10e9_L.s <- scale(chip_mat.phen.df$PLT_10e9_L)

chip_defined.df <- as.data.frame(chip_defined.mat)
chip_defined.df$eid <- as.double(rownames(chip_defined.df))
chip_mat.phen.df <- chip_mat.phen.df |> left_join(chip_defined.df, by = c("eid"))
```


```{r}
#x <- "thrombophlebitis"
#x <- "embolism"
#x <- "infarction"
AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "CHIP.sma", predictors = c("Sex", "Age", "BMI", "Smoking","thromb", "PLT_10e9_L.s","statin","antidiabetic","antithrombotic"))
AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "thromb", predictors = c("Sex", "Age", "BMI", "Smoking", "CHIP.all","PLT_10e9_L.s"))

AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "CHIP.all", predictors = c("Sex", "Age", "BMI", "Smoking","thromb"))
AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "thromb", predictors = c("Sex", "Age", "BMI", "Smoking", "CHIP.all"))

AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "statin", predictors = c("Sex", "Age", "BMI", "Smoking","CHIP.all","thromb"))
```
 

#### phen: self vs ukb
```{r}
pp <- list()
pp[[1]] <- (AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "union", predictors = c("Sex", "Age", "BMI", "Smoking")))$plot + ylim(0.95,1.25)
pp[[2]] <- (AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "self", predictors = c("Sex", "Age", "BMI", "Smoking")))$plot + ylim(0.95,1.25)
pp[[3]] <- (AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "ukb", predictors = c("Sex", "Age", "BMI", "Smoking")))$plot + ylim(0.95,1.25)
pp[[4]] <- (AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "self.only", predictors = c("Sex", "Age", "BMI", "Smoking")))$plot + ylim(0.95,1.25)

pdf("chip_f1g.pdf",width =4 , height =12,useDingbats = FALSE)
ggarrange(plotlist = pp, ncol = 1)
dev.off()
```

```{r}
chip.or <- list() 
chip.or$`Union CHIP` <-  (AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "union", predictors = c("Sex", "Age", "BMI", "Smoking")))$or
chip.or$`In-house CHIP` <-  (AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "self", predictors = c("Sex", "Age", "BMI", "Smoking")))$or
chip.or$`UKB-released CHIP` <-  (AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "ukb", predictors = c("Sex", "Age", "BMI", "Smoking")))$or
chip.or$`In-house only CHIP` <-  (AbelR::plot_logistic_forest(chip_mat.phen.df, outcome = "self.only", predictors = c("Sex", "Age", "BMI", "Smoking")))$or

chip.or2 <- lapply(names(chip.or),function(x){
  df <- chip.or[[x]]
  df$CHIP <- x
  return(df)
})
chip.or.df <- do.call(rbind,chip.or2)
write.table(chip.or.df,"chip_suppl_table_chipor.tsv",row.names = FALSE, quote = FALSE, sep = "\t")
```


```{r}
df_age <- list()
df_age[[1]] <- chip_mat.phen.df |> filter(union == 1) |> select(eid,Age) |> mutate(Type = "union") |> arrange(Age) |> mutate(ECD = row_number() / n())
df_age[[2]] <- chip_mat.phen.df |> filter(self == 1) |> select(eid,Age) |> mutate(Type = "self") |> arrange(Age) |> mutate(ECD = row_number() / n())
df_age[[3]] <- chip_mat.phen.df |> filter(self.only == 1) |> select(eid,Age) |> mutate(Type = "self.only") |> arrange(Age) |> mutate(ECD = row_number() / n())
df_age[[4]] <- chip_mat.phen.df |> filter(ukb == 1) |> select(eid,Age) |> mutate(Type = "ukb") |> arrange(Age) |> mutate(ECD = row_number() / n())
#df_age[[5]] <- chip_mat.phen.df |> filter(ukb == 1) |> select(eid,Age) |> mutate(Type = "overlap") |> arrange(Age) |> mutate(ECD = row_number() / n())
#df_age[[6]] <- chip_mat.phen.df |> filter(ukb == 1) |> select(eid,Age) |> mutate(Type = "ukb.only") |> arrange(Age) |> mutate(ECD = row_number() / n())
df_plot <- do.call(rbind,df_age)

unique(df_plot$Type)

p_ecd <- df_plot  |> ggplot( aes(x = Age, y = ECD, color = Type)) +
  geom_smooth(method = "loess",se = FALSE,span = 0.4,linewidth = 1) +
  scale_color_manual(values = alpha(c("union" = "black","self" = "#0098C9","self.only" = "#B5AA0F","overlap" = "#3C5488","ukb" = "#128640", "ukb.only" = "#748098"), 0.5)) +
  theme_test()

p_ecd
#pdf("chip_f1f.pdf",width = 8, height =6,useDingbats = FALSE)
#p_ecd
#dev.off()
```

```{r}
calc_ecd_cor <- function(df, ref_type = "In-house CHIP", by = 0.1) {
  
  age_grid <- seq(
    floor(min(df$Age, na.rm = TRUE)),
    ceiling(max(df$Age, na.rm = TRUE)),
    by = by
  )
  
  ecd_wide <- df %>%
    group_by(Type) %>%
    arrange(Age, .by_group = TRUE) %>%
    summarise(
      data = list(tibble(
        Age_grid = age_grid,
        ECD = approx(Age, ECD, xout = age_grid, rule = 2)$y
      )),
      .groups = "drop"
    ) %>%
    tidyr::unnest(data) %>%
    tidyr::pivot_wider(names_from = Type, values_from = ECD)
  
  other_types <- setdiff(names(ecd_wide), c("Age_grid", ref_type))
  
  tibble(
    Comparison = paste(ref_type, "vs", other_types),
    Pearson_r = sapply(other_types, function(x) {
      cor(ecd_wide[[ref_type]], ecd_wide[[x]], use = "complete.obs")
    })
  )
}

df_plot2 <- df_plot %>%
  mutate(
    Type = recode(
      Type,
      "union" = "Union CHIP",
      "self" = "In-house CHIP",
      "self.only" = "In-house-only CHIP",
      "ukb" = "UKB-released CHIP"
    )
  )

cor_df <- calc_ecd_cor(df_plot2, ref_type = "In-house CHIP")
cor_df
```

#### phen: CH large vs small
```{r}
CHIP <- names(eid.chip.list)
var.arr <- c("Sex","Age","BMI","Smoking","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")

tmp.or.list <- lapply(CHIP,function(x){
  df <- chip_mat.phen.df |> filter(CHIP.all == 0 | .data[[x]] ==1)
  or <- AbelR::plot_logistic_forest(df, outcome = x, predictors = var.arr)
  or$or$NUM <- sum(df[,x])
  or$or$Symbol <- str_split(x, "\\.")[[1]][1]
  or$or$Type <- str_split(x, "\\.")[[1]][2]
  return(or$or)
})
phen.or.df <- do.call(rbind,tmp.or.list)
write.table(phen.or.df,"chip_suppl_table_geneor.tsv",row.names = FALSE, sep = "\t", quote = FALSE)
```

```{r}
phen.or.df$Type <- factor(phen.or.df$Type, levels = rev(c("all","lar","sma")))
phen.or.df$Symbol <- factor(phen.or.df$Symbol, levels = c("CHIP","DNMT3A","TET2","ASXL1","PPM1D","TP53","SRSF2","JAK2","SF3B1"))
phen.or.df$Num <-paste0("N=",phen.or.df$NUM)
phen.or.df$Num <- factor(phen.or.df$Num, levels = rev(unique(phen.or.df$Num)))
phen.arr <- unique(phen.or.df$term)[c(1:4,15,16,17)]

plot.list <- lapply(phen.arr,\(x){
  phen.or.df |> filter(term  == x) |> ggplot(aes(x = Num, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange(aes(color = Type)) +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_text(aes(label = stars,color = Type), hjust = -1, show.legend = FALSE) + 
    coord_flip() +
    facet_grid(Symbol~.,scales = "free_y", switch = "y") +
    labs(y = "Odds Ratio (95% CI)", x = "", title = x) +
    theme_test() +  
    scale_color_manual(values = c("sma" = "#7BA39D", "lar" = "#B5AA0F", "all" = "#3C5488"))
})

plot.list

pdf("chip_f1e.pdf",width = 21, height =6,useDingbats = FALSE)
ggarrange(plotlist = plot.list, ncol = 7, common.legend = TRUE, legend = "top")
dev.off()
```

#### phen: thromb
```{r}
 var.arr <- c("Sex", "Age", "BMI", "Smoking", "pc1", "pc2", "pc3", "pc4", "pc5", "pc6", "pc7", "pc8", "pc9", "pc10", "statin", "antidiabetic", "antithrombotic")
#CHIP.arr <- names(eid.chip.list)[c(1:3,7:18)]
CHIP.arr <- names(eid.chip.list)
Throm.arr <- c("thromb","VTE","ATE") #"hyter",
#x = Throm.ar[1]
#y = CHIP.arr[1]
throm.or.ls <- lapply(Throm.arr,\(x){
  tmp.or.list <- lapply(CHIP.arr,\(y){
    df = chip_mat.phen.df |> filter(.data[[x]] == 1 |  thromb== 0) |> filter(CHIP.all == 0 | .data[[y]] ==1)
    or <- AbelR::plot_logistic_forest(df, outcome = x, predictors = c(y,var.arr))
    or$or$NUM <- dim(df |> filter(.data[[x]] == 1, .data[[y]] == 1))[1]
    or$or$Symbol <- str_split(y, "\\.")[[1]][1]
    or$or$Type <- str_split(y, "\\.")[[1]][2]
    or$or$Event <- x
    return(or$or[y,])
  })
  tmp.or.df <- do.call(rbind,tmp.or.list)
})
throm.or.df <- do.call(rbind,throm.or.ls)

write.table(throm.or.df,"chip_suppl_table_thtomor.tsv",sep = "\t", quote = FALSE, row.names = FALSE)
```
```{r}
events.colors <- c("black",mycol[1:8])
names(events.colors) <- c("CHIP","DNMT3A","TET2","ASXL1","PPM1D","TP53","SRSF2","JAK2","SF3B1")
throm.or.df$Type <- factor(throm.or.df$Type, levels = rev(c("all","lar","sma")))
throm.or.df$Symbol <- factor(throm.or.df$Symbol, levels = rev(c("CHIP","DNMT3A","TET2","ASXL1","PPM1D","JAK2","TP53","SRSF2","SF3B1")))
throm.or.df$Event <- factor(throm.or.df$Event, levels = Throm.arr)
throm.or.df$Num <-paste0("N=",throm.or.df$NUM)
  y_up <- max(throm.or.df$conf.high)
  y_dw <- min(throm.or.df$conf.low)
#throm.or.df$Num <- factor(throm.or.df$Num, levels = rev(unique(throm.or.df$Num)))

plot.list2 <- lapply(c("all","lar","sma"),\(x){

  df <- throm.or.df |> filter(Type == x)
  df$Type <- factor(paste(df$Event,df$Symbol), levels = paste(df$Event,levels(df$Symbol)))
  lab_vec <- setNames(df$Num, df$Type)
  df |> ggplot(aes(x = Type, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange(aes(color = Symbol)) +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_text(data = \(d) d |> dplyr::filter(!is.na(stars), stars != "ns"),
              aes(label = stars,color = Symbol), hjust = -1, show.legend = FALSE) + 
    scale_x_discrete(labels = lab_vec) + 
    scale_y_log10(limits = c(y_dw, y_up), breaks = c(0.5,  1, 3, 5),labels = c("0.5",  "1", "3",  "5")) + 
    coord_flip() +
    facet_grid(Event~.,scales = "free_y", switch = "y") +
    labs(y = "Odds Ratio (95% CI)", x = "", title = x) +
    theme_test() +  
    scale_color_manual(values = events.colors)
})

plot.list2

plot.list3 <- lapply(Throm.arr,\(x){
  throm.or.df |> filter(Event  == x) |> ggplot(aes(x = Num, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange(aes(color = Symbol)) +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_text(aes(label = stars,color = Symbol), hjust = -1, show.legend = FALSE) + 
    coord_flip() +
    facet_grid(Type~.,scales = "free_y", switch = "y") +
    labs(y = "Odds Ratio (95% CI)", x = "", title = x) +
    theme_test() +  
    scale_color_manual(values = rev(c("black",mycol[1:8])))
})

plot.list3

ggarrange(plotlist = plot.list2, ncol = 3, common.legend = TRUE, legend = "top")
```
```{r}
pdf("chip_f2d.pdf",width = 8, height =6,useDingbats = FALSE)
ggarrange(plotlist = plot.list2, ncol = 3, common.legend = TRUE, legend = "top")
dev.off()

```

####Distribution: thromb
```{r}
#ukb.icd10.thromb.df$eid <- as.double(ukb.icd10.thromb.df$eid)

#ukb.icd10.thromb.df |> filter(eid %in% chip_mat.phen.df$eid) |>  group_by(thromb_group,thromb_subgroup) |> count(thromb_group,thromb_subgroup) |> arrange(desc(n)) |>
#  ggplot(aes(y=thromb_subgroup,x=n)) + geom_col() + theme_test()
ukb.icd10.thromb.df <- ukb.icd10.thromb.df |>  filter(eid %in% chip_mat.phen.df$eid)
length(unique(ukb.icd10.thromb.df$eid))

df_group <- ukb.icd10.thromb.df |>
  count(thromb_group) |>
  filter(thromb_group %in% c("ATE", "VTE")) |>
  mutate(thromb_group = fct_reorder(thromb_group, -n))

p_all <- ggplot(df_group, aes(x = thromb_group, y = n, fill = thromb_group)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), vjust = -0.3, size = 5) +
  labs(x = NULL, y = "Number of ICD10 records") +
  theme_test() 

df_subgroup <- ukb.icd10.thromb.df |>
  filter(thromb_group %in% c("ATE", "VTE")) |>
  count(thromb_group, thromb_subgroup) |>
  group_by(thromb_group) |>
  mutate(thromb_subgroup = fct_reorder(thromb_subgroup, n)) |>
  ungroup()
df_subgroup$thromb_group <- factor(df_subgroup$thromb_group, levels = c("ATE", "VTE"))
p_sub <- ggplot(df_subgroup, aes(x = thromb_subgroup, y = n, fill = thromb_group)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n)), hjust = -0.1, size = 4) +
  coord_flip() +
  facet_grid(thromb_group~.,scales = "free_y", space = "free_y",switch = "y") +
  labs(x = NULL, y = "Number of ICD10 records") +
  theme_test() + ggtitle("Thrombosis events: N = 42,659") +
  scale_fill_manual(values = c("ATE" = "#00A087", "VTE" = "#4DBBD5")) + ylim(0,23000)

df_icd <- ukb.icd10.thromb.df |>
  count(ICD10, Diagnosis, thromb_group, thromb_subgroup,sort = TRUE) |>
  mutate(label = paste0(ICD10, ": ", Diagnosis),
         label = forcats::fct_reorder(label, n))



p_icd <- ggplot(df_icd, aes(x = label, y = n, fill = thromb_group)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Number of ICD10 records", fill = "Group") +
  theme_test(base_size = 12)

p_all
p_sub
p_icd

pdf("chip_f2e.pdf",width = 8, height = 4)
p_sub
dev.off()

write.table(df_icd,"chip_suppl_table_events.tsv",sep = "\t", quote = FALSE, row.names = FALSE)
```

### HR CHIP & disease
```{r}
#C92.Myeloid leukaemia
disease.icd10 <- "^C92" ##Myeloid leukaemia
disease_df.ml <- ukb.icd10.df |> filter(grepl(disease.icd10,ICD10)) |> filter(!duplicated(eid))
HR.ls.ml <- CHIP_disease_HR(disease_df.ml, "CHIP", "Myeloid leukaemia")

#C34.Primary lung cancer
disease.icd10 <- "^C34"
disease_df.plc <- ukb.icd10.df |> filter(grepl(disease.icd10,ICD10)) |> filter(!duplicated(eid))
HR.ls.plc <- CHIP_disease_HR(disease_df.plc, "CHIP", "Primary lung cancer")

#Thrombosis events
event_name <- "Thrombosis events"
disease_df.throm <- ukb.icd10.thromb.df |> filter(!duplicated(eid))
HR.ls.throm <- CHIP_disease_HR(disease_df.throm, "CHIP", "Thrombosis events")
```
```{r}
HR.ls.ml$pv
HR.ls.plc$pv
HR.ls.throm$pv

#pdf("chip_f2a.pdf", width = 7, height= 6 )
HR.ls.ml$plot
HR.ls.plc$plot
HR.ls.throm$plot
#dev.off()
```

```{r}
HR.throm.genes.ls <- lapply(c("DNMT3A","TET2","ASXL1","PPM1D","TP53","SRSF2","JAK2","SF3B1"), \(x){
  df <- CHIP_disease_HR(disease_df.throm, x, "Thrombosis events")
  return(df)
})
names(HR.throm.genes.ls) <- c("DNMT3A","TET2","ASXL1","PPM1D","TP53","SRSF2","JAK2","SF3B1")


#pdf("chip_f2b.pdf", width = 7, height= 6 )
HR.throm.genes.ls$DNMT3A$plot
HR.throm.genes.ls$TET2$plot
HR.throm.genes.ls$ASXL1$plot
HR.throm.genes.ls$PPM1D$plot
HR.throm.genes.ls$TP53$plot
HR.throm.genes.ls$SRSF2$plot
HR.throm.genes.ls$JAK2$plot
HR.throm.genes.ls$SF3B1$plot
#dev.off()
```

```{r}
CHIP_disease_HR <- function(disease_df, ch = "CHIP", event_name, suv_col = c("#AFC4E3", "#3C5488", "#C6DDD8", "#7BA39D", "#E6DE7A", "#B5AA0F")
) {
  
  ch_a <- paste0(ch, ".all")
  ch_s <- paste0(ch, ".sma")
  ch_l <- paste0(ch, ".lar")
  
  var.arr <- c("Sex", "Age", "BMI", "Smoking", "pc1", "pc2", "pc3", "pc4", "pc5", "pc6", "pc7", "pc8", "pc9", "pc10", "statin", "antidiabetic", "antithrombotic")
  
  disease_df <- disease_df |> inner_join(chip_mat.phen.df, by = "eid") |> 
    mutate( Sex = as.integer(Sex == "Male"), Smoking = as.integer(Smoking == "Yes") ) |>
    filter(.data[[ch_a]] == 1 | CHIP.all == 0) |> 
    mutate( CH_all = .data[[ch_a]], CH_small = .data[[ch_s]], CH_large = .data[[ch_l]], 
            time = if_else( status == 0, as.Date("2024-12-30") - date, Date_Death - date), 
            time_years = as.numeric(time) / 365.25)
  
  results <- list(fit = list(), plot = list(), pv = list())
  
  # Cox: all CHIP
  model <- coxph(reformulate(c(var.arr, "CH_all"), response = "Surv(time_years, status)"), data = disease_df)
  results$fit$cox_a <- model
  results$plot$cox_a <- ggforest(model,data = disease_df,main = paste0("Hazard ratio (", event_name, ")"))
  
  # Cox: small / large clone
  model <- coxph(reformulate(c(var.arr, "CH_small", "CH_large"), response = "Surv(time_years, status)"),data = disease_df)
  results$fit$cox_o <- model
  results$plot$cox_o <- ggforest(model,data = disease_df,main = paste0("Hazard ratio (", event_name, ")"))
  
  # KM: all CHIP
  fit <- survfit(Surv(time_years, status) ~ CH_all, data = disease_df)
  results$fit$suv_a <- fit
  
  sdiff <- survdiff(Surv(time_years, status) ~ CH_all, data = disease_df)
  results$pv$all <- pchisq(sdiff$chisq, df = length(sdiff$n) - 1, lower.tail = FALSE)
  p_lab <- paste0("Log-rank p = ", signif(results$pv$all, 3))
  
  results$plot$suv_a <- ggsurvplot(fit,data = disease_df,palette = suv_col[1:2],
                                   risk.table = TRUE,ncensor.plot = FALSE,pval = p_lab,
                                   conf.int = FALSE,xlab = "Time in years",
                                   title = paste0("Overall survival (", event_name, "): ", ch, " overall"),
                                   xlim = c(0, 20),break.time.by = 5,censor = TRUE,
                                   censor.shape = 124,censor.size = 1.5,size = 0.6)
  
  # KM: small clone
  disease_df_s <- disease_df |> filter(CH_small == 1 | CHIP.all == 0)
  
  fit <- survfit(Surv(time_years, status) ~ CH_small, data = disease_df_s)
  results$fit$suv_s <- fit
  
  sdiff <- survdiff(Surv(time_years, status) ~ CH_small, data = disease_df_s)
  results$pv$sma <- pchisq(sdiff$chisq, df = length(sdiff$n) - 1, lower.tail = FALSE)
  p_lab <- paste0("Log-rank p = ", signif(results$pv$sma, 3))
  
  results$plot$suv_s <- ggsurvplot(fit,data = disease_df_s,palette = suv_col[3:4],
                                   risk.table = TRUE,ncensor.plot = FALSE,pval = p_lab,
                                   conf.int = FALSE,xlab = "Time in years",
                                   title = paste0("Overall survival (", event_name, "): ", ch, " small"),
                                   xlim = c(0, 20),break.time.by = 5,censor = TRUE,
                                   censor.shape = 124,censor.size = 1.5,size = 0.6)
  
  # KM: large clone
  disease_df_l <- disease_df |>
    filter(CH_large == 1 | CHIP.all == 0)
  
  fit <- survfit(Surv(time_years, status) ~ CH_large, data = disease_df_l)
  results$fit$suv_l <- fit
  
  sdiff <- survdiff(Surv(time_years, status) ~ CH_large, data = disease_df_l)
  results$pv$lar <- pchisq(sdiff$chisq, df = length(sdiff$n) - 1, lower.tail = FALSE)
  p_lab <- paste0("Log-rank p = ", signif(results$pv$lar, 3))
  
  results$plot$suv_l <- ggsurvplot(fit,data = disease_df_l,palette = suv_col[5:6],
                                   risk.table = TRUE,ncensor.plot = FALSE,
                                   pval = p_lab,conf.int = FALSE,xlab = "Time in years",
                                   title = paste0("Overall survival (", event_name, "): ", ch, " large"),
                                   xlim = c(0, 20),break.time.by = 5,censor = TRUE,
                                   censor.shape = 124,censor.size = 1.5,size = 0.6)
  
  return(results)
}
```



### Blood biomarker odd ratio
```{r}
events <- c(colnames(ukb_blood_count_df)[!grepl("_pct$|eid$|_ratio$",colnames(ukb_blood_count_df))],
            colnames(ukb.biochem.df)[-1])
tmp.or.list <- lapply(events,function(x){
  #x  <- "PLT_10e9_L"
  #df <- chip_mat.phen.df |> filter(eid %in% unique(eid.ls$CBC,eid.ls$Biochem))
  df <- chip_mat.phen.df |> filter(eid %in% eid.ls$Biochem)
  df$event <- scale(df[,x])
  or <- AbelR::plot_logistic_forest(df, outcome = "CHIP.all", predictors = c(colnames(chip_mat.phen.df)[c(2,3,4,6,7:19)], "event"))
  or <- or$or["event",]
  rownames(or) <- paste0(x)
  return(or)
})

biomarker.or <- do.call(rbind,tmp.or.list)

biomarker.or$term <- rownames(biomarker.or)
biomarker.or$type <- c(rep("BloodCount",18),rep("BioChemistry",30))
biomarker.or <- biomarker.or |> arrange(estimate) |> mutate(term = factor(term, levels = term))
```
```{r}
biomarker.or <- biomarker.or |> mutate(label = paste0(round(estimate, 3), " (", round(conf.low, 3), "-", round(conf.high, 3), ") ",stars))
biomarker.or |> filter(type == "BloodCount") |> filter(p.value < 0.05) |> 
  ggplot( aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange(color = "firebrick") +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_text(aes(label = label), vjust = -1.2,  color = "black") + 
    coord_flip() +
    labs(y = "Odds Ratio (95% CI)", x = "", title = "Logistic regression CHIP") +
    #theme_minimal()
    theme_test() 

p_bm <- biomarker.or |> filter(type == "BioChemistry") |> filter(p.value < 0.05) |> 
  ggplot( aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange(color = "firebrick", size = 1.5) +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_text(aes(label = label), vjust = -1.2,  color = "black") + 
    coord_flip() +
    labs(y = "Odds Ratio (95% CI)", x = "", title = "Logistic regression CHIP") +
    #theme_minimal()
    theme_test() 
p_bm
pdf("chip_f4g.pdf",width = 8,height =6,useDingbats = FALSE)
p_bm
dev.off()
#write.table(biomarker.or,"ukb.biomarker.oddratio.txt",sep = "\t",row.names = FALSE)
```

### NMR meta & CHIP
```{r}
#events <- colnames(ukb_metabolic_df)[!grepl("(%)|eid$",colnames(ukb_metabolic_df))]
events <- colnames(ukb_metabolic_df)[!grepl("eid$",colnames(ukb_metabolic_df))]

tmp.or.list <- lapply(events,function(x){
  #x  <- "PLT_10e9_L"
  df <- chip_mat.phen.df |> inner_join(ukb_metabolic_df, by = c("eid"))
  df$event <- scale(log1p(df[,x]))
  or <- AbelR::plot_logistic_forest(df, outcome = "CHIP.all", predictors = c(colnames(chip_mat.phen.df)[c(2,3,4,6,7:16)], "event"))
  or <- or$or["event",]
  rownames(or) <- paste0(x)
  return(or)
})

nmr_meta.or <- do.call(rbind,tmp.or.list)
nmr_meta.or$term <- rownames(nmr_meta.or)
nmr_meta.or <- nmr_meta.or |> arrange(estimate) |> mutate(term = factor(term, levels = term))
```

```{r}
nmr_meta.or[c(1:5,247:251),] |> filter(p.value < 0.05) |> 
  ggplot( aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange(color = "firebrick") +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_text(aes(label = label), vjust = -1.2,  color = "black") + 
    coord_flip() +
    labs(y = "Odds Ratio (95% CI)", x = "", title = "Logistic regression CHIP") +
    #theme_minimal()
    theme_test() 
```
### olink & CHIP
```{r}
olink.df <- fread("olink_instance_0.NPX.csv")
keep_protein <- colnames(olink.df)[colMeans(!is.na(olink.df)) > 0.8]
keep_eid <- olink.df$eid[rowMeans(!is.na(olink.df)) > 0.8]
index_protein <- match(keep_protein,colnames(olink.df))
index_eid <- match(keep_eid, olink.df$eid)
olink.df <- olink.df[index_eid,..index_protein]
```
```{r}
events <- colnames(olink.df)[!grepl("eid$",colnames(olink.df))]

tmp.or.list <- lapply(events,function(x){
  #x  <- "fam20a"
  df <- chip_mat.phen.df |> inner_join(olink.df, by = c("eid"))
  df$event <- scale(df[,x])
  or <- AbelR::plot_logistic_forest(df, outcome = "CHIP.all", predictors = c(colnames(chip_mat.phen.df)[c(2,3,4,6,7:19)], "event"))
  or <- or$or["event",]
  rownames(or) <- paste0(x)
  return(or)
})

olink.or <- do.call(rbind,tmp.or.list)
olink.or$term <- rownames(olink.or)
olink.or <- olink.or |> arrange(estimate) |> mutate(term = factor(term, levels = term))
```

```{r}
olink.or[c(1:8,2903:2911),] |> filter(p.value < 0.05) |> 
  ggplot( aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_pointrange(color = "firebrick") +
    geom_hline(yintercept = 1, linetype = 2, color = "grey50") +
    geom_text(aes(label = label), vjust = -1.2,  color = "black") + 
    coord_flip() +
    labs(y = "Odds Ratio (95% CI)", x = "", title = "Logistic regression CHIP") +
    #theme_minimal()
    theme_test() 
```

#### GASE
```{r}
olink_anno.df <- fread("olink_resources/coding143.tsv") |> mutate(Protein = str_to_lower(str_replace(meaning, ";.*", ""))) |>  mutate(Symbol = str_replace(meaning, ";.*", ""))
olink_assay.df <- fread("olink_resources/olink_assay.dat") %>% mutate(Assay = tolower(Assay))
all(olink_anno.df$Protein %in% olink_assay.df$Assay)
olink_anno.df <- olink_anno.df |> left_join(olink_assay.df, by = c("Protein" = "Assay"))

#olink_anno.df$Protein
```

```{r}
olink.or[!olink.or$term %in% olink_anno.df$Protein,]$term <- c("hla-dra","hla-a","hla-e","ervv-1")
```

```{r}
olink.or$score <- sign(log2(olink.or$estimate)) * -log10(olink.or$p.value)
olink.or <- olink.or |>  arrange(desc(score)) %>% mutate(RANK = row_number())
olink.or <- olink.or |> left_join(olink_anno.df, by = c("term" = "Protein"))
olink.or$Entrez <- TransGeneID(olink.or$Symbol, "Symbol", "Entrez", organism = "hsa")
olink.or[is.na(olink.or$Entrez),]
```

```{r}
  deg.df <- olink.or |> filter(! is.na(Entrez))
  library(msigdbr)
  m_t2g.H <- msigdbr(species = "Homo sapiens", category = "H") %>% dplyr::select(gs_name, entrez_gene)
  gene.fc <- deg.df$score
  names(gene.fc) <- deg.df$Entrez
  gene.fc<-sort(gene.fc, decreasing = TRUE)
  gsea.H <- GSEA(gene.fc, TERM2GENE = m_t2g.H, pvalueCutoff = 1)
  gsea.H <- setReadable(gsea.H, org.Hs.eg.db, keyType = "ENTREZID")
  gsea.kegg <-  gseKEGG(gene.fc, organism = "hsa", keyType = "kegg", pvalueCutoff = 1)
  gsea.kegg <- setReadable(gsea.kegg, org.Hs.eg.db, keyType = "ENTREZID")
  gsea.go_bp <- gseGO(gene.fc, org.Hs.eg.db, ont = "BP", pvalueCutoff = 1, keyType = "ENTREZID")
  gsea.go_bp <- setReadable(gsea.go_bp, org.Hs.eg.db, keyType = "ENTREZID")
```


```{r}
plot.list1 <- lapply(c("hsa05202", "hsa05221", "hsa04672", "hsa04919"),function(x){
  gsea_plot_custorm(gsea.kegg,x,"#7BA39D",xpos = 1000)
})

plot.list2 <- lapply(c("hsa04151", "hsa04979", "hsa03320", "hsa04060"),function(x){
  gsea_plot_custorm(gsea.kegg,x,"#3C5488",xpos = 1000)
})

plot_grid(plotlist = plot.list1)
plot_grid(plotlist = plot.list2)
```

```{r}
gsea_plot_custorm(gsea.go_bp,"GO:0008015","#B5AA0F",xpos = 1000)
gsea_plot_custorm(gsea.go_bp,"GO:0035295","#B5AA0F",xpos = 1000)

gsea.plot <- gsea.go_bp@result |> select(Description, NES, pvalue, p.adjust) |>  top_n(n=40, wt = -pvalue)

gsea.plot$Group <- "Up"
gsea.plot[gsea.plot$NES < 0,]$Group <- "Down"
gsea.plot$Lable <- "GO_BP"

ggplot(gsea.plot, aes(x = NES, y = Description, fill = Group)) + geom_col() + theme_test() + geom_vline(xintercept = 0)

ggplot(gsea.plot, aes(x = Lable, y = fct_reorder(Description, -log10(pvalue)))) + geom_point(aes(size = -log10(pvalue), color = NES)) + theme_test() + scale_color_gradient2(low = "#08306B",mid = "grey90", high = "#67000D")
```

```{r}
iogo <- "GO:0008015"
iogo <- "GO:0035295"
target_genes <- gsea.go_bp@result |> filter(ID ==iogo) |> pull(core_enrichment) |> strsplit("/") |> unlist()
target_desc <- gsea.go_bp@result |> filter(ID ==iogo) |> pull(Description)

  library("ggrastr")
  gg <- olink.or
  iof <- target_genes
  gg$label <- ""
  gg[match(iof,gg$Symbol),]$label <- iof
  p <- ggplot(gg, aes(x = RANK/1000, y= score))
  p <- p + geom_point_rast(color = "grey", shape = 16, size = 2, alpha = 0.6, show.legend = FALSE)
  p <- p + geom_point(color = ifelse(gg$label == "" ,NA,"#B5AA0F"), shape = 16, size = 2, alpha = 0.6, show.legend = FALSE)
  p <- p + geom_point(color = ifelse(gg$label == "" ,NA,"black"), shape = 1, size = 2, show.legend = FALSE)
  p = p + ggrepel::geom_text_repel(aes(label=label),color="#B5AA0F",show.legend = FALSE,fontface = "bold", size = 2.5,box.padding = unit(0.8, "lines"),point.padding = unit(0.3, "lines"), segment.size = 0.3, max.overlaps = 8000)
  p = p + geom_hline(yintercept = 0, linetype = "dotted")
  p + theme_test()  + ggtitle(target_desc) + ylab("Score")
```

#### AML vs non-Cancer
```{r}
olink.df$eid
disease.icd10 <- "^C92" ##Myeloid leukaemia

ukb.icd10.df

aml_df <- ukb.icd10.df |> filter(grepl("^C92.0|^C92.5|^C92.6|^C92.8",ICD10)) |> filter(!duplicated(eid))
cancer_df <- ukb.icd10.df |> filter(Category == "Neoplasms (Tumors)") |> filter(!duplicated(eid))
aml_olink.eid <- olink.df$eid[olink.df$eid %in% aml_df$eid] 
non_olink.eid <- olink.df$eid[!olink.df$eid %in% cancer_df$eid] 
#non_olink.eid <- olink.df$eid[!olink.df$eid %in% ukb.icd10.df$eid] 


aml.ogt <- olink.df |> filter(eid %in% aml_olink.eid) |> select(ogt) |> filter(! is.na(ogt))
non.ogt <- olink.df |> filter(eid %in% non_olink.eid) |> select(ogt) |> filter(! is.na(ogt))


aml.ogt$Group <- "AML"
non.ogt$Group <- "non-Cancer"

mean(aml.ogt$ogt)
mean(non.ogt$ogt)
```

```{r}
plot.df <- rbind(aml.ogt,non.ogt)

ggplot(plot.df, aes(y = ogt, x = Group)) + geom_violin(aes(fill = Group)) + theme_test() + geom_boxplot(width=0.1) +
  scale_fill_manual(values = c("#3C5488FF", "#00A087FF")) + 
  stat_compare_means(method = "wilcox.test",paired = FALSE, label = "p.signif",
                       comparisons=list(c("AML", "non-Cancer")))
```

###Competing risks

#### dataframe
```{r}
lag_days <- 180
event_name <- "Thrombosis events"

competing_risk.df <- ukb.icd10.thromb.df |> group_by(eid) |> slice_min(order_by = date) |> ungroup() |> filter(!duplicated(eid))
competing_risk.df <- competing_risk.df |> right_join(chip_mat.phen.df,by = "eid")
competing_risk.df <- competing_risk.df |> mutate(Sex = as.integer(Sex == "Male"),Smoking = as.integer(Smoking == "Yes"))
competing_risk.df$assess_date <- competing_risk.df$Date

competing_risk.df <- competing_risk.df |> filter(is.na(date) | date > (assess_date + lag_days)) |>
  mutate(end_date = pmin(date,as.Date("2024-12-31"),Date_Death, na.rm = TRUE),
         time = as.numeric(end_date - assess_date),
         status = case_when(!is.na(date) & date <= end_date ~ 1,
                            is.na(date) & !is.na(Date_Death) & Date_Death <= as.Date("2024-12-31") ~ 2,
                            TRUE ~ 0
                            )
         )

competing_risk.df |> group_by(status, CHIP.all) |> summarise(n())
competing_risk.df$status <- as.factor(competing_risk.df$status)

mycol3 <- c("#AFC4E3","#3C5488","#C6DDD8","#7BA39D","#E6DE7A","#B5AA0F","#E88FB5","#C81E76")
```

```{r}
competing_risk.result <- list()
competing_risk.result$throm <- competing_risk_plot(competing_risk.df,"CHIP.all","Thrombosis event")
#competing_risk.result$hyper <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "Hypercoagulability"),"CHIP.all","Hypercoagulability")
competing_risk.result$ate <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "ATE"),"CHIP.all","ATE")
competing_risk.result$vte <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "VTE"),"CHIP.all","VTE")
```
```{r}
pdf("chip_f2c.pdf", width = 5 , height = 5, useDingbats = FALSE)
competing_risk.result$throm$cr_plot
competing_risk.result$hyper$cr_plot
competing_risk.result$ate$cr_plot 
competing_risk.result$vte$cr_plot
dev.off()
```

```{r}
competing_risk.result$vte.tet2 <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "VTE") |> filter(CHIP.all == 0 |  TET2.all== 1),"TET2.all","VTE")
competing_risk.result$vte.ppm1d <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "VTE") |> filter(CHIP.all == 0 |  PPM1D.all== 1),"PPM1D.all","VTE")
competing_risk.result$ate.tet2 <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "ATE") |> filter(CHIP.all == 0 |  TET2.all== 1),"TET2.all","ATE")
competing_risk.result$ate.ppm1d <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "ATE") |> filter(CHIP.all == 0 |  PPM1D.all== 1),"PPM1D.all","ATE")
```
```{r}
competing_risk.result$vte.dnmt3a <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "VTE") |> filter(CHIP.all == 0 |  DNMT3A.all== 1),"DNMT3A.all","VTE")
competing_risk.result$vte.asxl1 <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "VTE") |> filter(CHIP.all == 0 |  ASXL1.all== 1),"ASXL1.all","VTE")
competing_risk.result$ate.dnmt3a <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "ATE") |> filter(CHIP.all == 0 |  DNMT3A.all== 1),"DNMT3A.all","ATE")
competing_risk.result$ate.asxl1 <- competing_risk_plot(competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "ATE") |> filter(CHIP.all == 0 |  ASXL1.all== 1),"ASXL1.all","ATE")
```

```{r}
competing_risk.result$vte.dnmt3a$cr_plot
competing_risk.result$vte.asxl1$cr_plot
competing_risk.result$vte.ppm1d$cr_plot
competing_risk.result$vte.tet2$cr_plot

competing_risk.result$ate.dnmt3a$cr_plot
competing_risk.result$ate.asxl1$cr_plot
competing_risk.result$ate.ppm1d$cr_plot
competing_risk.result$ate.tet2$cr_plot
```
```{r}

ch <- c("DNMT3A.all","ASXL1.all","TET2.all","PPM1D.all")
var.arr = c("Sex","Age","BMI","Smoking","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")
   df <- competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "VTE") |> filter(CHIP.all == 0 |  DNMT3A.all== 1 | ASXL1.all == 1 | TET2.all == 1 | PPM1D.all == 1 )
  fg <- finegray(reformulate(c(ch, var.arr,"eid"), response = "Surv(time/365.25, status)"), data=df, etype = 1)
  fit_fg_adj <- coxph(reformulate(c(ch,var.arr), response = "Surv(fgstart, fgstop, fgstatus)"),data = fg, weights = fgwt, ties = "efron", robust = TRUE,cluster = eid)
  ggforest(fit_fg_adj, data = fg, main = "sHR (Fine-Gray Model)")
  
  vte.fg.sHR.df <- do.call(rbind,lapply(ch,function(x){extract_coxph(fit_fg_adj,x,"VTE")}))
  
  df <- competing_risk.df |> filter(is.na(thromb_group) | thromb_group == "ATE") |> filter(CHIP.all == 0 |  DNMT3A.all== 1 | ASXL1.all == 1 | TET2.all == 1 | PPM1D.all == 1 )
  fg <- finegray(reformulate(c(ch, var.arr,"eid"), response = "Surv(time/365.25, status)"), data=df, etype = 1)
  fit_fg_adj <- coxph(reformulate(c(ch,var.arr), response = "Surv(fgstart, fgstop, fgstatus)"),data = fg, weights = fgwt, ties = "efron", robust = TRUE,cluster = eid)
  ggforest(fit_fg_adj, data = fg, main = "sHR (Fine-Gray Model)")
  ate.fg.sHR.df <- do.call(rbind,lapply(ch,function(x){extract_coxph(fit_fg_adj,x,"ATE")}))
```
```{r}
df_plot <- rbind(ate.fg.sHR.df,vte.fg.sHR.df)
df_plot <- df_plot |> dplyr::mutate(p_lab = case_when(p < 0.0001 ~ "****", p < 0.001 ~ "***", p < 0.01  ~ "**", p < 0.05  ~ "*", TRUE ~ "ns"))
df_plot <- df_plot |> dplyr::mutate(label = paste0(round(HR, 2), " [", round(lower, 2), "-", round(upper, 2), "] ",p_lab))
df_plot$Model <- factor(df_plot$Model, levels = rev(c("VTE", "ATE")))
df_plot$Variable <- factor(df_plot$Variable, levels = c("DNMT3A.all","TET2.all","ASXL1.all","PPM1D.all"))
p_shr <- ggplot(df_plot, aes(y = Model, x = HR)) + 
    geom_point(aes(color = Model),size = 5,show.legend = FALSE, shape = 16) +
    geom_errorbarh(aes(xmin = lower, xmax = upper,color = Model), height = 0.1,show.legend = FALSE) +
    geom_vline(xintercept = 1, linetype = 2, color = "grey50") +
    scale_color_manual(values =c("#4DBBD5","#00A087")) + 
    geom_text(aes(x = HR , label = label), vjust = -1,show.legend = FALSE) + 
    theme_test() + ylab("Fine–Gray model") + facet_grid(Variable~.,scales = "free_y", switch = "y") + 
    xlab("Subdistribution Hazard Ratio (sHR)")

p_shr
 
pdf("chip_f2f.pdf",width = 5 , height = 5, useDingbats = FALSE)
p_shr
dev.off()
```

#### function
```{r}
extract_coxph <- function(fit, var_name, label){
  s <- summary(fit)$coefficients
  data.frame(
    Model = label,
    Variable = var_name,
    HR = s[var_name, "exp(coef)"],
    lower = exp(s[var_name, "coef"] - 1.96 * s[var_name, "se(coef)"]),
    upper = exp(s[var_name, "coef"] + 1.96 * s[var_name, "se(coef)"]),
    p = s[var_name, "Pr(>|z|)"]
  )
}

competing_risk_plot <- function(df, ch, event, var.arr = c("Sex","Age","BMI","Smoking","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10")){
  library(survival)
#  df = competing_risk.df 
 # ch = "CHIP.all"
  #event = "Thrombosis event"
  #var.arr <- c("Sex","Age","BMI","Smoking","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10")
  
  mycol3 <- c("#AFC4E3","#3C5488","#C6DDD8","#7BA39D","#E6DE7A","#B5AA0F")
  result = list()
  ############fine-gray model 
  fg <- finegray(reformulate(c(ch, var.arr,"eid"), response = "Surv(time/365.25, status)"), 
               data=df, etype = 1)
  fit_fg_crude <- coxph(reformulate(c(ch), response = "Surv(fgstart, fgstop, fgstatus)"),
                     data = fg, weights = fgwt, ties = "efron", robust = TRUE,cluster = eid)
  fit_fg_adj <- coxph(reformulate(c(ch,var.arr), response = "Surv(fgstart, fgstop, fgstatus)"),
                     data = fg, weights = fgwt, ties = "efron", robust = TRUE,cluster = eid)
  result$adj <- ggforest(fit_fg_adj, data = fg, main = "sHR (Fine-Gray Model)")
  result$unadj <- ggforest(fit_fg_crude, data = fg, main = "sHR (Fine-Gray Model)")
  df_plot <- bind_rows(extract_coxph(fit_fg_crude, ch, "Unadjusted"), extract_coxph(fit_fg_adj, ch, "Adjusted"))
  df_plot <- df_plot |> dplyr::mutate(p_lab = case_when(p < 0.0001 ~ "****", p < 0.001 ~ "***", p < 0.01  ~ "**", p < 0.05  ~ "*", TRUE ~ "ns"))
  df_plot <- df_plot |> dplyr::mutate(label = paste0(round(HR, 2), " [", round(lower, 2), "-", round(upper, 2), "] ",p_lab))
  df_plot$Model <- factor(df_plot$Model, levels = c("Unadjusted", "Adjusted"))
  p1 <- ggplot(df_plot, aes(y = Model, x = HR)) + 
    geom_point(aes(color = Model),size = 4,show.legend = FALSE) +
    geom_errorbarh(aes(xmin = lower, xmax = upper,color = Model), height = 0.1,show.legend = FALSE) +
    geom_vline(xintercept = 1, linetype = 2, color = "grey50") +
    scale_x_log10() + scale_color_manual(values = mycol3[1:2]) + 
    geom_text(aes(x = HR , label = label), vjust = -1,show.legend = FALSE) + 
    theme_test() + ylab("Fine–Gray model") + 
    xlab("Subdistribution Hazard Ratio (sHR)") 
#(Death as competing event)

################ CIF (Cumulative incidence function)
  library(cmprsk)
  ci <- cuminc(ftime = df$time/365.25, fstatus = df$status, group = df[[ch]]) 
  result$cif <- ci

  df_cif <- rbind(
    #data.frame(time=ci[["0 2"]]$time, est=ci[["0 2"]]$est, grp="Death CHIP=0"),
    #data.frame(time=ci[["1 2"]]$time, est=ci[["1 2"]]$est, grp="Death CHIP=1"),
    data.frame(time=ci[["0 1"]]$time, est=ci[["0 1"]]$est, grp="CHIP-"),
    data.frame(time=ci[["1 1"]]$time, est=ci[["1 1"]]$est, grp="CHIP+")
  )
  pval_cif <- ci$Tests[1, "pv"]
  lab <- ifelse(pval_cif < 0.001,"Gray's test P < 0.001",paste0("Gray's test P = ", sprintf("%.3f", pval_cif)))
  p2 <- ggplot(df_cif, aes(time, est, color=grp)) +
    geom_step(size=1) +
    labs(x="Years", y="Cumulative incidence", color="Group") +
    annotate("text", x = 10, y = max(df_cif$est) * 0.25, label = lab, hjust = 0) +
    theme_test() + theme(legend.position = "inside", legend.position.inside = c(0.15,0.75)) +
    scale_color_manual(values = mycol3[1:2])   

  result$cr_plot <- p2 / p1 + plot_layout(heights = c(2.5, 1))  + 
    plot_annotation(title = paste0(event," (exposure = ", ch,")"), 
                    theme = theme(plot.title = element_text(hjust = 0.5,face = "bold"))) 
  return(result)
}
```

## 04. CHIP vs Olink
### phenotype
```{r}
olink.df <- fread("olink_instance_0.NPX.csv")
keep_protein <- colnames(olink.df)[colMeans(!is.na(olink.df)) > 0.8]
keep_eid <- olink.df$eid[rowMeans(!is.na(olink.df)) > 0.8]
index_protein <- match(keep_protein,colnames(olink.df))
index_eid <- match(keep_eid, olink.df$eid)
olink.df <- olink.df[index_eid,..index_protein]

protein.arr <- setdiff(colnames(olink.df), "eid")

ukb_wes_all_eid <- eid.ls$Olink
chip_mut_df <- data.frame(eid = ukb.maf.df$Tumor_Sample_Barcode, CHIP = paste0(ukb.maf.df$Hugo_Symbol,":",ukb.maf.df$AAChange))
olink_wes.df <- olink.df |> filter(eid %in% ukb_wes_all_eid) |> left_join(chip_mat.phen.df, by = c("eid"))
```

```{r}
olink_wes.df |> group_by(CHIP.all) |> summarise(Num = n())
olink_wes.df <- as.data.frame(olink_wes.df)
```

```{r}
ggplot(olink_wes.df, aes(x = CHIP.all, y = Age)) + geom_violin(aes(fill = CHIP.all)) + geom_boxplot(width  = 0.25) + theme_test() 
ggplot(olink_wes.df, aes(x = Sex, y = Age)) + geom_violin(aes(fill = CHIP.all)) + theme_test() 
ggplot(olink_wes.df, aes(x = Sex, y = Age)) + geom_violin(aes(fill = Sex)) + geom_boxplot(width  = 0.25) + theme_test() 
ggplot(olink_wes.df, aes(x = CHIP.all, y = BMI)) + geom_violin(aes(fill = CHIP.all)) + geom_boxplot(width  = 0.25) + theme_test() 
```

### limma diff
```{r}
library(limma)
var.arr <- c("Age", "Sex", "Smoking","BMI","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")
expr.df <- olink_wes.df[complete.cases(olink_wes.df[, var.arr]), ]
design <- model.matrix(reformulate(c("CHIP.all",var.arr)), expr.df)

expr <- expr.df[,protein.arr]
rownames(expr) <- expr.df$eid
#expr <- as.matrix(expr)
#storage.mode(expr) <- "numeric"

fit <- lmFit(t(expr), design)
fit <- eBayes(fit)
chip.olink.limma_df <- topTable(fit, coef="CHIP.all",number = Inf)
```

```{r}
chip.olink.limma_df["a1bg",]
fit$coefficients
fit$stdev.unscaled * fit$sigma
```

```{r}
olink_anno.df <- fread("olink_resources/coding143.tsv") |> mutate(Protein = str_to_lower(str_replace(meaning, ";.*", ""))) |>  mutate(Symbol = str_replace(meaning, ";.*", ""))
olink_assay.df <- fread("olink_resources/olink_assay.dat") %>% mutate(Assay = tolower(Assay))
all(olink_anno.df$Protein %in% olink_assay.df$Assay)
olink_anno.df <- olink_anno.df |> left_join(olink_assay.df, by = c("Protein" = "Assay"))
```
```{r}
p_olink <- limma_olink_volcano(chip.olink.limma_df,top=10, title = "Olink CH+ vs CH-")
p_olink + theme(legend.position = "top")
```

```{r}
pdf("chip_f4c.pdf",width = 6,height = 8)
p_olink + theme(legend.position = "top")
dev.off()
```


### GASE
```{r}
idx <- !rownames(chip.olink.limma_df) %in% olink_anno.df$Protein
rownames(chip.olink.limma_df)[idx]
sum(idx)
rownames(chip.olink.limma_df)[idx] <- c("hla-dra", "ervv-1", "hla-e", "hla-a")
sum(!rownames(chip.olink.limma_df) %in% olink_anno.df$Protein)
```

```{r}
olink.lm <- chip.olink.limma_df
#olink.lm$score <- sign(olink.lm$logFC) * -log10(olink.lm$P.Value)
olink.lm$score <- olink.lm$t
olink.lm <- olink.lm |>  arrange(desc(score)) %>% mutate(RANK = row_number())
olink.lm$Protein <- rownames(olink.lm) 
olink.lm <- olink.lm |> left_join(olink_anno.df, by = c("Protein"))
olink.lm$Entrez <- TransGeneID(olink.lm$Symbol, "Symbol", "Entrez", organism = "hsa")
olink.lm[is.na(olink.lm$Entrez),]
```

```{r}
  deg.df <- olink.lm |> filter(! is.na(Entrez))
  library(msigdbr)
  m_t2g.H <- msigdbr(species = "Homo sapiens", category = "H") %>% dplyr::select(gs_name, entrez_gene)
  gene.fc <- deg.df$score
  names(gene.fc) <- deg.df$Entrez
  gene.fc<-sort(gene.fc, decreasing = TRUE)
  gsea.H <- GSEA(gene.fc, TERM2GENE = m_t2g.H, pvalueCutoff = 1)
  gsea.H <- setReadable(gsea.H, org.Hs.eg.db, keyType = "ENTREZID")
  gsea.kegg <-  gseKEGG(gene.fc, organism = "hsa", keyType = "kegg", pvalueCutoff = 1)
  gsea.kegg <- setReadable(gsea.kegg, org.Hs.eg.db, keyType = "ENTREZID")
  gsea.go_bp <- gseGO(gene.fc, org.Hs.eg.db, ont = "BP", pvalueCutoff = 1, keyType = "ENTREZID")
  gsea.go_bp <- setReadable(gsea.go_bp, org.Hs.eg.db, keyType = "ENTREZID")
  
  
write.table(gsea.go_bp@result |> filter(pvalue < 0.05), "chip_suppl_table_olinkgsea.tsv", row.names = FALSE, quote = FALSE, sep = "\t")
```

```{r}
plot.list1 <- lapply(c("hsa05202", "hsa05221", "hsa04672", "hsa04919"),function(x){
  gsea_plot_custorm(gsea.kegg,x,"#7BA39D",xpos = 1000)
})

plot.list2 <- lapply(c("hsa04151", "hsa04979", "hsa03320", "hsa04060"),function(x){
  gsea_plot_custorm(gsea.kegg,x,"#3C5488",xpos = 1000)
})

plot_grid(plotlist = plot.list1)
plot_grid(plotlist = plot.list2)
```

```{r}
plot.list1 <- lapply(c("GO:0006325", "GO:0002250", "GO:0045088"),function(x){
  gsea_plot_custorm(gsea.go_bp,x,"#B5AA0F",xpos = 1000)
})

plot.list2 <- lapply(c("GO:0097529", "GO:0060326", "GO:0008015"),function(x){
  gsea_plot_custorm(gsea.go_bp,x,"#3C5488",xpos = 1000)
})

plot.list1
plot.list2
plot_grid(plotlist = c(plot.list1,plot.list2),nrow = 2)
```
```{r}
pdf("chip_f4e.pdf",width = 10,height = 6)
plot_grid(plotlist = c(plot.list1,plot.list2),nrow = 2)
dev.off()
```

```{r}
iof.gobp <- c(
  "tube development",
  "blood vessel development",
  "angiogenesis",
  "myeloid leukocyte migration",
  "granulocyte migration",
  "chromatin organization",
  "cell chemotaxis",
  "circulatory system process",
  "blood circulation",
  "chromosome segregation",
  "chromatin remodeling",
  "chromosome organization",
  "adaptive immune response",
  "mononuclear cell migration",
  "regulation of immune response",
  "regulation of innate immune response",
  "DNA damage response",
  "negative regulation of B cell proliferation"
)

#gsea.plot <- gsea.go_bp@result |> select(Description, NES, pvalue, p.adjust) |>  top_n(n=40, wt = -p.adjust)
gsea.plot <- gsea.go_bp@result |> select(Description, NES, pvalue, p.adjust) |>  filter(Description %in% iof.gobp)

gsea.plot$Group <- "Up"
gsea.plot[gsea.plot$NES < 0,]$Group <- "Down"
gsea.plot$Lable <- "GO_BP"

#ggplot(gsea.plot, aes(x = NES, y = Description, fill = Group)) + geom_col() + theme_test() + geom_vline(xintercept = 0)

p_go <- ggplot(gsea.plot, aes(x = Lable, y = fct_reorder(Description, -log10(p.adjust)))) + geom_point(aes(size = -log10(p.adjust), color = NES)) + theme_test() + scale_color_gradient2(low = "#08306B",mid = "grey90", high = "#67000D")
p_go
```

```{r}
pdf("chip_f4d.pdf",width = 4.5,height = 4)
p_go
dev.off()
```
```{r}
p_core <- plot_gsea_core_genes(gsea.go_bp@result, "GO:0045088",c("CD14", "TLR1", "TLR2", "CD274", "VSIG4", "LILRA2", "LILRB1", "CLEC6A", "CLEC7A", "LAMP2", "TRIM21", "TRIM25"),color = "#B5AA0F") + plot_gsea_core_genes(gsea.go_bp@result, "GO:0060326",c("CCL2", "CCL5", "CCL8", "CCL14", "CCL18", "CCL21", "CCL24", "CXCL9", "CXCL10", "CXCL16", "CX3CL1", "VEGFA", "VEGFC", "SERPINE1", "PPBP"),color = "#3C5488")
p_core
```

```{r}
pdf("chip_f4f.pdf",width = 6,height = 4)
p_core
dev.off()
```

```{r}
#iogo <- "GO:0008015"
#iogo <- "GO:0002250"

plot_gsea_core_genes <- function(gsea.res,iogo,label,color = "#B5AA0F"){
  target_genes <- gsea.res|> filter(ID ==iogo) |> pull(core_enrichment) |> strsplit("/") |> unlist()
  target_desc <-gsea.res |> filter(ID ==iogo) |> pull(Description)
  library("ggrastr")
  gg <- olink.lm
  iof <-target_genes
  gg$iof <- ""
  gg[match(iof,gg$Symbol),]$iof <- iof
  gg$label <- ""
  gg[match(label,gg$Symbol),]$label <- label
  p <- ggplot(gg, aes(x = RANK, y= score))
  p <- p + geom_point_rast(color = "grey", shape = 16, size = 2, alpha = 0.6, show.legend = FALSE)
  p <- p + geom_point(color = ifelse(gg$iof == "" ,NA,color), shape = 16, size = 2, alpha = 0.6, show.legend = FALSE)
  p <- p + geom_point(color = ifelse(gg$iof == "" ,NA,"black"), shape = 1, size = 2, show.legend = FALSE)
  p = p + ggrepel::geom_text_repel(aes(label=label),color=color,show.legend = FALSE,fontface = "bold", size = 2.5,box.padding = unit(1, "lines"),point.padding = unit(0.3, "lines"), segment.size = 0.3, max.overlaps = 2000)
  p = p + geom_hline(yintercept = 0, linetype = "dotted")
  p = p + theme_test()  + ggtitle(target_desc) + ylab("Moderated t statistic") + xlab("Protein rank")
  return(p)
}
```


## 05. CHIP vs blood count
### combind CHIP
```{r}
cbc_traits.arr <- setdiff(colnames(ukb_blood_count_df), "eid")
ukb_wes_all_eid <- eid.ls$CBC
ukb_blood_count_df.chip <- chip_mat.phen.df |> filter(eid %in% ukb_wes_all_eid)
```

### limma diff
```{r}
library(limma)
var.arr <- c("Age", "Sex", "Smoking","BMI","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")
expr.df <- ukb_blood_count_df.chip[complete.cases(ukb_blood_count_df.chip[, var.arr]), ]
design <- model.matrix(reformulate(c("CHIP.all",var.arr)), expr.df)
expr <- scale(expr.df[,cbc_traits.arr])
rownames(expr) <- expr.df$eid

fit <- lmFit(t(expr), design)
fit <- eBayes(fit)
chip.bloodcount.limma_df <- topTable(fit, coef="CHIP.all",number = Inf)
```


```{r}
limma_bloodcount_volcano(limma_df = chip.bloodcount.limma_df, title = "CH+ vs CH-")
```
```{r}
ggplot(ukb_blood_count_df.chip, aes(x = CHIP, y = log2(PLT_10e9_L))) + geom_violin(aes(fill = CHIP)) + geom_boxplot(width  = 0.25) + theme_test() 

ggplot(ukb_blood_count_df.chip, aes(x = CHIP, y = PCT_pct)) + geom_violin(aes(fill = CHIP)) + geom_boxplot(width  = 0.25) + theme_test()

ukb_blood_count_df.chip |> group_by(CHIP) |> summarise(mean(PLT_10e9_L))
```

```{r}
p_cbc_1 <- limma_bloodcount_volcano(limma_df = chip.bloodcount.limma_df, title = "CH+ vs CH-")
p_cbc_2 <- ggplot(ukb_blood_count_df.chip, aes(x = as.factor(CHIP.all), y = log2(PLT_10e9_L))) + geom_violin(aes(fill = as.factor(CHIP.all))) + geom_boxplot(width  = 0.25) + theme_test() 
p_cbc_1 + p_cbc_2 + plot_layout(widths = c(2:1))

pdf("chip_f4b.pdf", width = 6, height = 3.5)
p_cbc_1 + p_cbc_2 + plot_layout(widths = c(2:1))
dev.off()
```

```{r}
ggplot(ukb_blood_count_df.chip, aes(x = as.factor(CHIP.all), y = log2(Neutrophils_10e9_L))) + geom_violin(aes(fill = as.factor(CHIP.all))) + geom_boxplot(width  = 0.25) + theme_test()
```

###biochem limma
```{r}
biochem_traits.arr <- setdiff(colnames(ukb.biochem.df), "eid")
ukb_wes_all_eid <- eid.ls$Biochem
ukb.biochem.df.chip <- chip_mat.phen.df |> filter(eid %in% ukb_wes_all_eid)


library(limma)
var.arr <- c("Age", "Sex", "Smoking","BMI","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")
expr.df <- ukb.biochem.df.chip[complete.cases(ukb.biochem.df.chip[, var.arr]), ]
design <- model.matrix(reformulate(c("CHIP.all",var.arr)), expr.df)
expr <- scale(expr.df[,biochem_traits.arr])
rownames(expr) <- expr.df$eid

fit <- lmFit(t(expr), design)
fit <- eBayes(fit)
chip.biochem.limma_df <- topTable(fit, coef="CHIP.all",number = Inf)

limma_metabolic_volcano(limma_df = chip.biochem.limma_df, title = "CH+ vs CH-")
```


## 06. NMR metabolomics

### combind CHIP
```{r}
nmr_traits.arr <- setdiff(colnames(ukb_metabolic_df), "eid")
ukb_wes_all_eid <- eid.ls$NMR
ukb_metabolic_df.chip <- ukb_metabolic_df |> filter(eid %in% ukb_wes_all_eid) |> left_join(chip_mat.phen.df, by = c("eid"))
```

### limma diff
```{r}
library(limma)
var.arr <- c("Age", "Sex", "Smoking","BMI","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")
expr.df <- ukb_metabolic_df.chip[complete.cases(ukb_metabolic_df.chip[, var.arr]), ]

##########all 

design <- model.matrix(reformulate(c("CHIP.all",var.arr)), expr.df)
expr <- scale(expr.df[,nmr_traits.arr])
rownames(expr) <- expr.df$eid

fit <- lmFit(t(expr), design)
fit <- eBayes(fit)
chip.metabolic.limma_df <- topTable(fit, coef="CHIP.all",number = Inf)

##########percentatge
meta.per <- colnames(expr.df)[grepl("%",colnames(expr.df))]
expr.per <- scale(expr.df[,meta.per])
rownames(expr.per) <- expr.df$eid
fit <- lmFit(t(expr.per), design)
fit <- eBayes(fit)
chip.metabolic.per.limma_df <- topTable(fit, coef="CHIP.all",number = Inf)

########## mmol/l
meta.mml <- colnames(expr.df)[grepl("mmol/l",colnames(expr.df))]
expr.mml <- scale(expr.df[,meta.mml])
rownames(expr.mml) <- expr.df$eid
fit <- lmFit(t(expr.mml), design)
fit <- eBayes(fit)
chip.metabolic.mml.limma_df <- topTable(fit, coef="CHIP.all",number = Inf)

```
```{r}
limma_metabolic_volcano(chip.metabolic.limma_df, top =10, title = "CH+ vs CH-")
limma_metabolic_volcano(chip.metabolic.per.limma_df, top =10, title = "CH+ vs CH-")
limma_metabolic_volcano(chip.metabolic.mml.limma_df, top =10, title = "CH+ vs CH-")

pdf("chip_f4h.pdf", width = 4, height = 4.2)
limma_metabolic_volcano(chip.metabolic.mml.limma_df, top =10, title = "CH+ vs CH-")
dev.off()
```


```{r}
write.table(chip.bloodcount.limma_df,"chip_suppl_table_limmaCBC.tsv",quote = FALSE, sep = "\t")
write.table(chip.biochem.limma_df,"chip_suppl_table_limmaBioC.tsv",quote = FALSE, sep = "\t")
write.table(chip.metabolic.limma_df,"chip_suppl_table_limmanNMR.tsv",quote = FALSE, sep = "\t")
write.table(olink.lm,"chip_suppl_table_limmaOlink.tsv",quote = FALSE, sep = "\t")
```

## 07. GWAS
### QC
```{r}
#######input
ukb.gwas_pheno.df <- read.csv("GWAS_data/ukb.gwas_pheno.csv", na.strings = c("", "NA"))
ukb.gwas_pheno.df <- ukb.gwas_pheno.df |> 
  rename_with(~ gsub("participant.","", .x)) |> 
  rename_with(~ gsub("p22009_a","pc", .x)) |> 
  rename(sex = p31, age = p21022, ever_smoked = p20160_i0, bmi = p23104_i0,
         hdl_cholesterol = p30760_i0, ldl_cholesterol = p30780_i0,
         genetic_sex = p22001, ethnic = p22006, aneuploidy = p22019) |> 
  select(-c(p22418,p21008,p22021))
########filter
ukb.gwas_pheno.df |> count(p22020,ethnic,aneuploidy, sex == genetic_sex)
ukb.gwas_pheno.df <- ukb.gwas_pheno.df |> filter(sex == genetic_sex) |> filter(ethnic == 1) |>  filter(p22020 == 1) |> filter(is.na(aneuploidy))
########hypertension
ukb.gwas_pheno.df <- ukb.gwas_pheno.df |> 
  mutate( hypertension = as.integer(rowSums(across(all_of(c("p2966_i0","p2966_i1","p2966_i2","p2966_i3")), ~ as.numeric(.)), na.rm = TRUE) > 0) ) |> 
  select(-c(p2966_i0, p2966_i1, p2966_i2, p2966_i3, p22020, aneuploidy, genetic_sex, ethnic))
########fill na
ukb.gwas_pheno.df <- ukb.gwas_pheno.df |>
  mutate(ever_smoked = if_else(is.na(ever_smoked), 0, ever_smoked)) |> 
  mutate(across(all_of(c("bmi","hdl_cholesterol","ldl_cholesterol")), ~ if_else(is.na(.),mean(., na.rm = TRUE), .))) 

####### thromb_event 
ukb.gwas_pheno.df <- ukb.gwas_pheno.df |> 
  left_join(ukb.icd10.thromb.df[,c("eid","thromb_group")], by = c("eid") ) |> 
  rename(thromb_event = thromb_group) |>
  mutate(thromb_cc = if_else(is.na(thromb_event),0,1))

####### CHIP mut
ukb.gwas_pheno.df <- ukb.gwas_pheno.df |> inner_join(chip_mat.phen.df[,c("eid","CHIP.all")], by = c("eid"))
ukb.gwas_pheno.df |> count(thromb_cc, CHIP.all)
```
```{r}
ukb_21008.samples <- read.table("GWAS_data/ukb21008_c1_b0_v1.list", sep = "", head = TRUE)
all(ukb.gwas_pheno.df$eid %in% ukb_21008.samples$ID_2)
```
```{r}
ukb.gwas_pheno.df$IID <- ukb.gwas_pheno.df$eid
ukb.gwas_pheno.df$FID <- ukb.gwas_pheno.df$eid
ukb.gwas_pheno.df <- ukb.gwas_pheno.df[,c("FID","IID",colnames(ukb.gwas_pheno.df)[2:21])]

sum(ukb.gwas_pheno.df$CHIP.all)
sum(ukb.gwas_pheno.df$thromb_cc)
```

### PheWAS
#### exsample
```{r}
#devtools::install_github("PheWAS/PheWAS")
library("PheWAS")
set.seed(1)
ex=generateExample()
id.vocab.code.count=ex$id.vocab.code.count
genotypes=ex$genotypes
id.sex=ex$id.sex

id.vocab.code.count
genotypes
id.sex

phenotypes=createPhenotypes(id.vocab.code.count, 
  aggregate.fun=sum, id.sex=id.sex)
data=inner_join(inner_join(phenotypes,genotypes),id.sex)

results=phewas_ext(data,phenotypes=names(phenotypes)[-1],genotypes=c("rsEXAMPLE"),
  covariates=c("sex"), cores=1)

phewasManhattan(results, title="My Example PheWAS Manhattan Plot")
```
#### CHIP
```{r}
ukb.gwas_pheno.df <- read.csv("GWAS_data/ukb.gwas_pheno.chip_thromb.tsv", sep = "\t")

id.vocab.code.count <- ukb.icd10.df |> filter(eid %in% ukb.gwas_pheno.df$IID) |> select(eid,ICD10)  |> mutate(vocabulary_id="ICD10", count =1) |> rename(id=eid,code=ICD10) 
id.vocab.code.count <- id.vocab.code.count[,c("id","vocabulary_id","code","count")]
id.sex <- ukb.gwas_pheno.df |> select(IID,sex) |> mutate(sex = if_else(sex == 1, "M", "F")) |> rename(id=IID)
genotypes <- ukb.gwas_pheno.df |> select(IID,CHIP.all) |> rename(id=IID, withCHIP = CHIP.all)

phenotypes = createPhenotypes(id.vocab.code.count, min.code.count=1, add.phecode.exclusions=T, translate=T, vocabulary.map=PheWAS::phecode_map_icd10, id.sex = id.sex)

covariate <- ukb.gwas_pheno.df |> rename(id=IID) |>
  select(- c(FID,hdl_cholesterol,ldl_cholesterol,
             hypertension,hypertension,thromb_event,thromb_cc,CHIP.all))

results=phewas(phenotypes, genotypes, cores = 10, 
               significance.threshold = c('p-value', 'bonferroni', 'fdr'),
               covariates = covariate, additive.genotypes = FALSE)
```
```{r}
library("PheWAS")
load("GWAS_data/PheWAS.RData")

results <- results |> mutate(p = if_else( p< 1e-35 & !is.na(p),1e-35,p))
phewasManhattan(results, title="CHIPmut PheWAS Manhattan Plot")

results_d <- addPhecodeInfo(results)
res = results_d[results_d$bonferroni & !is.na(results_d$p),]

```

## 08.Combind

### list
```{r}
#CBC Biochem WES Olink NMR

remove_rows_by_na_ratio <- function(df, cols, threshold = 0.8) {
  # df: data.frame
  # cols: 指定列范围，比如 2:31，或 c("A", "B", "C")
  # threshold: NA占比阈值，默认0.8
  na_ratio <- rowMeans(is.na(df[, cols, drop = FALSE]))
  df_filtered <- df[na_ratio <= threshold, , drop = FALSE]
  return(df_filtered)
}

wes_eids <- as.character(chip_mat.phen.df$eid)

eid.ls <- list()
eid.ls$WES <- wes_eids
eid.ls$CBC <- intersect(as.character(remove_rows_by_na_ratio(ukb_blood_count_df, 2:32)$eid), wes_eids)
eid.ls$Biochem <- intersect(as.character(remove_rows_by_na_ratio(ukb.biochem.df, 2:31)$eid), wes_eids)
eid.ls$Olink <- intersect(as.character(olink.df$eid), wes_eids)
eid.ls$NMR <- intersect(as.character(remove_rows_by_na_ratio(ukb_metabolic_df, 2:252)$eid), wes_eids)
```
```{r}
omics_summary <- lapply(names(eid.ls), function(nm) {
  tmp <- chip_mat.phen.df[chip_mat.phen.df$eid %in% eid.ls[[nm]], ]

  data.frame(
    Dataset = nm,
    Total = nrow(tmp),
    CHIP_pos = sum(tmp$CHIP.all == 1, na.rm = TRUE),
    CHIP_neg = sum(tmp$CHIP.all == 0, na.rm = TRUE)
  )
}) |> 
  dplyr::bind_rows()

omics_summary
```

### upset
```{r}
library(ComplexHeatmap)
design.mat <- list_to_matrix(eid.ls)
design.comb.mat <- make_comb_mat(design.mat)

p_upset <- UpSet(design.comb.mat, 
      set_order = c("WES","CBC","Biochem","NMR","Olink"),
      comb_col = c("#5f75ae", "#7ba39d", "#de8e06","#6a3d9a","#A81E2C")[comb_degree(design.comb.mat)],  
      bg_col = "#F0F0FF", bg_pt_col = "#CCCCFF", 
      top_annotation = upset_top_annotation(design.comb.mat, add_numbers = TRUE,
                                            annotation_name_rot = 90),
      right_annotation = upset_right_annotation(design.comb.mat, add_numbers = TRUE,
                                                show_annotation_name = FALSE,
                                                axis_param = list(side = "top"),
                                                gp = gpar(fill = "#08537C", col = "#08537C")
                                                )
      )
p_upset
```
```{r}
pdf("chip_f4a.pdf", width = 6, height = 3.5)
p_upset
dev.off()
```

```{r}
chip.olink.limma_df |> filter(adj.P.Val < 0.05)
chip.bloodcount.limma_df |>  filter(adj.P.Val < 0.05)
chip.biochem.limma_df |>  filter(adj.P.Val < 0.05)
chip.metabolic.mml.limma_df |>  filter(adj.P.Val < 0.05)
```
### extract data
```{r}
combind_eid <- rownames(design.mat)[rowSums(design.mat) == 5]
combind_dat <- chip_mat.phen.df |> filter(eid %in% combind_eid)
combind_dat <- combind_dat[,c(1:48,111:113)]
```

```{r}
olink_sign <- c(
  "flt3lg", "kitlg", "csf3",
  "cd1c", "clec4c", "lamp3", "cd86",
  "sell", "ccl5",
  "ceacam8", "clec4d", "cd5l", "ager", "cd36", "cd300lf", "cd209",
  "angpt1", "vegfc", "axl", "tnfsf12",
  "il32", "crisp3", "prg3", "lama4", "ccn2", "ddah1", "lpl", "lag3", "gzma", "klrb1"
)

cbc_sign <- c("PLT_10e9_L","Neutrophils_10e9_L","Monocytes_10e9_L","Eosinophils_10e9_L","Lymphocytes_10e9_L","RBC_10e12_L")

biochem_sign <- c("Cystatin_C(mg/L)","Cholesterol(mmol/L)","LDL_direct(mmol/L)","HDL_cholesterol(mmol/L)","Apolipoprotein_B(g/L)","Apolipoprotein_A(g/L)","C-reactive_protein(mg/L)","Gamma_glutamyltransferase(U/L)","Urate(umol/L)")

nmr_sign <- c("Linoleic Acid (mmol/l)","Omega-6 Fatty Acids (mmol/l)","Polyunsaturated Fatty Acids (mmol/l)","Docosahexaenoic Acid (mmol/l)","LDL Cholesterol (mmol/l)","HDL Cholesterol (mmol/l)","Total Cholesterol (mmol/l)","Acetone (mmol/l)")

length(c(olink_sign,cbc_sign,biochem_sign,nmr_sign))
```

```{r}
olink_dat <- olink_wes.df[,c("eid",olink_sign)] |> filter(eid %in% combind_eid)
nmr_dat <- ukb_metabolic_df.chip[,c("eid",nmr_sign)] |> filter(eid %in% combind_eid)
cbc_dat <- ukb_blood_count_df.chip[,c("eid",cbc_sign)] |> filter(eid %in% combind_eid)  
biochem_dat <- ukb.biochem.df.chip[,c("eid",biochem_sign)] |> filter(eid %in% combind_eid) 

combind_dat <- combind_dat |> left_join(olink_dat, by = "eid") 
combind_dat <- combind_dat |> left_join(nmr_dat, by = "eid")
combind_dat <- combind_dat |> left_join(cbc_dat, by = "eid")
combind_dat <- combind_dat |> left_join(biochem_dat, by = "eid")
```

### scoring
```{r}

library(dplyr)

winsorize_vec <- function(x, probs = c(0.005, 0.995)) {
  qs <- quantile(x, probs = probs, na.rm = TRUE)
  x[x < qs[1]] <- qs[1]
  x[x > qs[2]] <- qs[2]
  x
}

prep_z_matrix <- function(dat, features, log_transform = TRUE, winsor = TRUE, impute_na = TRUE) {
  features_use <- intersect(features, colnames(dat))
  
  if (length(features_use) == 0) {
    stop("No selected features found in dat.")
  }
  
  missing_features <- setdiff(features, colnames(dat))
  if (length(missing_features) > 0) {
    message("Missing features: ", paste(missing_features, collapse = ", "))
  }
  
  mat <- dat[, features_use, drop = FALSE]
  mat <- as.data.frame(lapply(mat, function(x) as.numeric(as.character(x))))
  
  if (log_transform) {
    mat <- as.data.frame(lapply(mat, function(x) {
      if (any(x < 0, na.rm = TRUE)) {
        return(x)
      } else {
        return(log1p(x))
      }
    }))
  }
  
  if (winsor) {
    mat <- as.data.frame(lapply(mat, winsorize_vec))
  }
  
  zmat <- as.data.frame(scale(mat))
  colnames(zmat) <- features_use
  
  if (impute_na) {
    zmat[] <- lapply(zmat, function(x) {
      x[is.na(x) | is.nan(x) | is.infinite(x)] <- 0
      x
    })
  }
  
  return(zmat)
}


get_beta <- function(limma_df, features, beta_col = "logFC") {
  common <- intersect(features, rownames(limma_df))
  
  if (length(common) == 0) {
    stop("No selected features found in limma_df rownames.")
  }
  
  missing_beta <- setdiff(features, rownames(limma_df))
  if (length(missing_beta) > 0) {
    message("Features not found in limma result: ", paste(missing_beta, collapse = ", "))
  }
  
  beta <- limma_df[common, beta_col]
  names(beta) <- common
  beta <- beta[!is.na(beta)]
  beta
}

calc_beta_score <- function(zmat, beta) {
  common <- intersect(colnames(zmat), names(beta))
  
  if (length(common) == 0) {
    stop("No common features between zmat and beta.")
  }
  
  zmat_use <- as.matrix(zmat[, common, drop = FALSE])
  beta_use <- beta[common]
  
  score <- as.numeric(zmat_use %*% beta_use) / sum(abs(beta_use))
  
  return(score)
}
```
```{r}
olink_z   <- prep_z_matrix(combind_dat, olink_sign, log_transform = FALSE, winsor = TRUE, impute_na = TRUE)
cbc_z     <- prep_z_matrix(combind_dat, cbc_sign, log_transform = TRUE, winsor = TRUE, impute_na = TRUE)
biochem_z <- prep_z_matrix(combind_dat, biochem_sign, log_transform = TRUE, winsor = TRUE, impute_na = TRUE)
nmr_z     <- prep_z_matrix(combind_dat, nmr_sign, log_transform = TRUE, winsor = TRUE,impute_na = TRUE)

beta_olink   <- get_beta(chip.olink.limma_df, olink_sign)
beta_cbc     <- get_beta(chip.bloodcount.limma_df, cbc_sign)
beta_biochem <- get_beta(chip.biochem.limma_df, biochem_sign)
beta_nmr     <- get_beta(chip.metabolic.mml.limma_df, nmr_sign)

combind_dat$score_olink   <- calc_beta_score(olink_z, beta_olink)
combind_dat$score_cbc     <- calc_beta_score(cbc_z, beta_cbc)
combind_dat$score_biochem <- calc_beta_score(biochem_z, beta_biochem)
combind_dat$score_nmr     <- calc_beta_score(nmr_z, beta_nmr)
```

```{r}
score_mat <- combind_dat[, c("score_olink", "score_cbc", "score_biochem", "score_nmr")]
score_mat_z <- as.data.frame(scale(score_mat))
colnames(score_mat_z) <- c("score_olink_z", "score_cbc_z", "score_biochem_z", "score_nmr_z")
combind_dat$score_multiomic <- rowMeans(score_mat_z, na.rm = FALSE)
combind_dat$score_multiomic_z <- as.numeric(scale(combind_dat$score_multiomic))
combind_dat<- cbind(combind_dat,score_mat_z)


combind_dat |> group_by(CHIP.all) |> summarise(median(score_multiomic_z),mean(score_multiomic_z))
```


```{r}
combind_dat |> ggplot(aes(x=as.factor(CHIP.all),y = score_olink)) + geom_boxplot()
combind_dat |> ggplot(aes(x=as.factor(CHIP.all),y = score_cbc)) + geom_boxplot()
combind_dat |> ggplot(aes(x=as.factor(CHIP.all),y = score_biochem)) + geom_boxplot()
combind_dat |> ggplot(aes(x=as.factor(CHIP.all),y = score_multiomic)) + geom_boxplot()
```
### heatmap
```{r}
feature_module_df <- dplyr::bind_rows(
  data.frame(feature = cbc_sign,     module = "CBC"),
  data.frame(feature = olink_sign,   module = "Olink"),
  data.frame(feature = biochem_sign, module = "Biochem"),
  data.frame(feature = nmr_sign,     module = "NMR")
)

feature_module_df$feature <- as.character(feature_module_df$feature)
feature_module_df$module <- factor(
  feature_module_df$module,
  levels = c("CBC", "Olink", "Biochem", "NMR")
)


feature_z <- cbind(
  cbc_z[, intersect(cbc_sign, colnames(cbc_z)), drop = FALSE],
  olink_z[, intersect(olink_sign, colnames(olink_z)), drop = FALSE],
  biochem_z[, intersect(biochem_sign, colnames(biochem_z)), drop = FALSE],
  nmr_z[, intersect(nmr_sign, colnames(nmr_z)), drop = FALSE]
)

feature_z <- as.data.frame(feature_z)
feature_z$eid <- combind_dat$eid

```

```{r}
get_beta_df <- function(limma_df, features, module, beta_col = "logFC") {
  common <- intersect(features, rownames(limma_df))
  
  data.frame(
    feature = common,
    beta = limma_df[common, beta_col],
    module = module,
    stringsAsFactors = FALSE
  )
}

beta_df <- dplyr::bind_rows(
  get_beta_df(chip.bloodcount.limma_df, cbc_sign, "CBC"),
  get_beta_df(chip.olink.limma_df, olink_sign, "Olink"),
  get_beta_df(chip.biochem.limma_df, biochem_sign, "Biochem"),
  get_beta_df(chip.metabolic.mml.limma_df, nmr_sign, "NMR")
)
```

```{r}
row_anno_df <- feature_module_df |>
  dplyr::left_join(beta_df |> dplyr::select(feature, beta), by = "feature")

row_anno_df <- row_anno_df |>
  dplyr::arrange(module, dplyr::desc(beta))

feature_order <- row_anno_df$feature
```

```{r}
set.seed(123)

idx_chip_neg <- which(combind_dat$CHIP.all == 0)
idx_chip_pos <- which(combind_dat$CHIP.all == 1)

idx_use <- c(
  sample(idx_chip_neg, min(4000, length(idx_chip_neg))),
  sample(idx_chip_pos, min(1000, length(idx_chip_pos)))
)

dat_hm <- combind_dat[idx_use, ]
mat_hm <- feature_z[idx_use, feature_order, drop = FALSE]
```

```{r}
mat_hm <- t(as.matrix(mat_hm))
mat_hm[mat_hm > 2] <- 2
mat_hm[mat_hm < -2] <- -2
sample_order <- order(
  dat_hm$CHIP.all,
  dat_hm$score_multiomic_z
)

dat_hm <- dat_hm[sample_order, ]
mat_hm <- mat_hm[, sample_order, drop = FALSE]
```

```{r}
library(ComplexHeatmap)
library(circlize)
library(grid)

row_anno_plot <- row_anno_df |>
  dplyr::filter(feature %in% rownames(mat_hm)) |>
  dplyr::distinct(feature, module, beta)

row_anno_plot <- row_anno_plot[match(rownames(mat_hm), row_anno_plot$feature), ]

row_ha <- rowAnnotation(
  Module = row_anno_plot$module,
  `CHIP β` = anno_barplot(
    row_anno_plot$beta,
    gp = gpar(fill = ifelse(row_anno_plot$beta >= 0, "#A81E2C", "#3C5488")),
    border = FALSE,
    width = unit(1.5, "cm")
  ),
  col = list(
    Module = c(
      CBC = "#0098C9",
      Olink = "#E64B35",
      Biochem = "#cab2d6",
      NMR = "#00A087"
    )
  )
)

library(zoo)

score_line <- zoo::rollmean(
  dat_hm$score_multiomic_z,
  k = 25,
  fill = NA,
  align = "center"
)

score_group_mean <- ave(
  dat_hm$score_multiomic_z,
  dat_hm$CHIP.all,
  FUN = function(x) mean(x, na.rm = TRUE)
)

top_ha <- HeatmapAnnotation(
  CHIP = factor(dat_hm$CHIP.all, levels = c(0, 1), labels = c("CHIP-", "CHIP+")),
  Score = anno_lines(
    score_line,
    gp = gpar(col = "black", lwd = 1),
    ylim = range(dat_hm$score_multiomic_z, na.rm = TRUE),
    height = unit(1.2, "cm")
  ),
  col = list(
    CHIP = c("CHIP-" = "#727d94", "CHIP+" = "#bb216f")
  ),
  annotation_name_side = "left"
)

col_fun <- colorRamp2(
  c(-2, 0, 2),
  c("#3C5488", "white", "#A81E2C")
)

#pdf("chip_f5a.pdf", width = 10, height = 7)
Heatmap(
  mat_hm,
  name = "Z-score",
  col = col_fun,
  top_annotation = top_ha,
  left_annotation = row_ha,
  row_split = row_anno_plot$module,
  column_split = factor(dat_hm$CHIP.all, levels = c(0, 1), labels = c("CHIP-", "CHIP+")),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_column_names = FALSE,
  show_row_names = TRUE,
  row_names_gp = gpar(fontsize = 4),
  column_title = NULL,
  heatmap_legend_param = list(
    title = "Feature\nz-score",
    legend_height = unit(4, "cm")
  )
)
#dev.off()


```

```{r}
row_anno_export <- row_anno_plot |>
  mutate(
    feature = if_else(module == "Olink", toupper(feature), feature),
    Direction = if_else(beta > 0, "Higher in CHIP", "Lower in CHIP")
  ) |>
  rename(
    `Feature ID` = feature,
    `Data layer` = module,
    `CHIP-associated beta / score weight` = beta
  )

row_anno_export

write.table(row_anno_export,"chip_suppl_table_mlscore.tsv",row.names = FALSE, sep = "\t", quote = FALSE)
```

### multi score
```{r}
var.arr <- c("Age", "Sex", "Smoking","BMI","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")

score_cols <- c(
  "score_cbc_z",
  "score_olink_z",
  "score_biochem_z",
  "score_nmr_z",
  "score_multiomic_z"
)

get_score_beta <- function(score_col, dat, covar,event) {
  
  form <- reformulate(
    termlabels = c(event, covar),
    response = score_col
  )
  
  fit <- lm(form, data = dat)
  
  broom::tidy(fit, conf.int = TRUE) |>
    dplyr::filter(term == event) |>
    dplyr::mutate(score = score_col)
}


score_beta_df <- dplyr::bind_rows(
  lapply(score_cols, get_score_beta, dat = combind_dat, covar = var.arr, event = "CHIP.all")
)
score_beta_df2 <- dplyr::bind_rows(
  lapply(score_cols, get_score_beta, dat = combind_dat, covar = var.arr, event = "thromb")
)

score_beta_df <- rbind(score_beta_df, score_beta_df2)

score_beta_df$score <- dplyr::recode(
  score_beta_df$score,
  score_cbc_z = "CBC",
  score_olink_z = "Olink",
  score_biochem_z = "Biochem",
  score_nmr_z = "NMR",
  score_multiomic_z = "Integrated"
)

score_beta_df$score <- factor(score_beta_df$score, levels = rev(c("Integrated","Olink","CBC","Biochem","NMR")))

p_f5_1 <- ggplot(score_beta_df, aes(x = estimate, y = score)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.15) +
  geom_point(size = 2.5) +
  theme_classic() +labs( x = "Adjusted β", y = NULL) +
  facet_wrap(~term )
p_f5_1
```

### driver-specific
```{r}
combind_dat <- combind_dat |> mutate(
  driver_group = case_when(
    CHIP.all == 0 ~ "CHIP-",
    DNMT3A.all == 1 ~ "DNMT3A",
    TET2.all == 1 ~ "TET2",
    ASXL1.all == 1 ~"ASXL1",
    PPM1D.all == 1 ~ "PPM1D",
    CHIP.all == 1 ~ "Others",
    TRUE ~ NA_character_
  ),
  thrombosis = case_when(
    thromb == 0 ~ "Thromb-",
    VTE == 1 ~ "VTE",
    ATE == 1 ~ "ATE",
    TRUE ~ NA_character_
  ),
  vaf_group = case_when(
    CHIP.all == 0 ~ "CHIP-",
    CHIP.sma == 1 ~ "Small_clone",
    CHIP.lar == 1 ~ "Large_clone",
    TRUE ~ NA_character_
  ),
  score_q = ntile(score_multiomic_z, 4),
  score_q = factor(score_q,levels = 1:4,labels = c("Q1", "Q2", "Q3", "Q4"))
)

combind_dat$driver_group <- factor(combind_dat$driver_group,levels = c("CHIP-", "DNMT3A", "TET2", "ASXL1", "PPM1D",  "Others"))
combind_dat$thrombosis <- factor(combind_dat$thrombosis,levels = c("Thromb-", "VTE", "ATE"))
```
```{r}
  event <- "score_multiomic_z"
  var.arr <- c("Age", "Sex", "Smoking","BMI","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")
  
  fit <- lm(reformulate(c("driver_group",var.arr), response = event), data = combind_dat)
  driver_score_df <- broom::tidy(fit, conf.int = TRUE) |>
  dplyr::filter(grepl("^driver_group", term)) |>
  dplyr::mutate(
    driver = gsub("^driver_group", "", term),
    driver = factor(driver, levels = c("DNMT3A", "TET2", "ASXL1", "PPM1D", "Others"))
  )

p_f5_2 <- ggplot(driver_score_df, aes(x = estimate, y = driver)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.18, linewidth = 0.7) +
  geom_point(size = 2.6) +
  theme_classic(base_size = 13) +
  labs(
    x = "Adjusted β for integrated multi-omic score",
    y = NULL
  )
p_f5_2 
```
```{r}
  event <- "score_multiomic_z"
  var.arr <- c("Age", "Sex", "Smoking","BMI","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")
  
  fit <- lm(reformulate(c("vaf_group",var.arr), response = event), data = combind_dat)
  driver_score_df <- broom::tidy(fit, conf.int = TRUE) |>
  dplyr::filter(grepl("^vaf_group", term)) |>
  dplyr::mutate(
    vaf = gsub("^vaf_group", "", term)
  )

ggplot(driver_score_df, aes(x = estimate, y = vaf)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.18, linewidth = 0.7) +
  geom_point(size = 2.6) +
  theme_classic(base_size = 13) +
  labs(
    x = "Adjusted β for integrated multi-omic score",
    y = NULL
  )

driver_score_df
```

```{r}
fit_thromb_q <- glm(
  reformulate(c("score_q", var.arr), response = "thromb"),
  data = combind_dat,
  family = binomial()
)

thromb_q_df <- broom::tidy(fit_thromb_q, conf.int = TRUE, exponentiate = TRUE) |>
  filter(grepl("^score_q", term)) |>
  mutate(
    group = gsub("score_q", "", term),
    group = factor(group, levels = c("Q2", "Q3", "Q4"))
  )

thromb_q_df
```


### score quantile
```{r}
library(dplyr)
library(ggplot2)
library(broom)
library(forcats)

dat_thromb <- combind_dat |>
  mutate(
    CHIP.all = as.integer(CHIP.all),
    thromb = as.integer(thromb),
    score_multiomic_z = as.numeric(score_multiomic_z)
  ) |>
  filter(
    !is.na(thromb),
    !is.na(CHIP.all),
    !is.na(score_multiomic_z)
  )

# 确认 thromb 是 0/1
table(dat_thromb$thromb, useNA = "ifany")
table(dat_thromb$CHIP.all, useNA = "ifany")
summary(dat_thromb$score_multiomic_z)

dat_thromb <- dat_thromb |>
  mutate(
    Sex = as.factor(Sex),
    Smoking = as.factor(Smoking)
  )

dat_thromb <- dat_thromb |>
  mutate(
    score_q = ntile(score_multiomic_z, 4),
    score_q = factor(
      score_q,
      levels = 1:4,
      labels = c("Q1", "Q2", "Q3", "Q4")
    )
  )

table(dat_thromb$score_q, dat_thromb$thromb)

fit_thromb_q <- glm(
  reformulate(c("score_q", var.arr), response = "thromb"),
  data = dat_thromb,
  family = binomial()
)

thromb_q_df <- broom::tidy(fit_thromb_q, conf.int = TRUE, exponentiate = TRUE) |>
  filter(grepl("^score_q", term)) |>
  mutate(
    group = gsub("score_q", "", term),
    group = factor(group, levels = c("Q2", "Q3", "Q4"))
  )

thromb_q_df

thromb_q_plot <- bind_rows(
  data.frame(
    term = "score_qQ1",
    estimate = 1,
    conf.low = 1,
    conf.high = 1,
    p.value = NA_real_,
    group = factor("Q1", levels = c("Q1", "Q2", "Q3", "Q4"))
  ),
  thromb_q_df |>
    mutate(group = factor(as.character(group), levels = c("Q1", "Q2", "Q3", "Q4")))
)


dat_thromb <- dat_thromb |>
  mutate(score_q_num = as.numeric(score_q))

fit_thromb_trend <- glm(
  reformulate(c("score_q_num", var.arr), response = "thromb"),
  data = dat_thromb,
  family = binomial()
)

trend_p <- broom::tidy(fit_thromb_trend, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term == "score_q_num")

trend_p

p5E <- ggplot(thromb_q_plot, aes(x = estimate, y = group)) +
  geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.18, linewidth = 0.7) +
  geom_point(size = 2.6) +
  scale_x_log10() +
  theme_classic(base_size = 13) +
  labs(
    x = "Adjusted OR for thromb",
    y = "Multi-omic score quartile"
  ) +
  annotate(
    "text",
    x = max(thromb_q_plot$conf.high, na.rm = TRUE),
    y = 1,
    hjust = 1,
    label = paste0("P trend = ", signif(trend_p$p.value, 3)),
    size = 4
  )

p5E


```
### CHIP + score
```{r}
q_score <- quantile(dat_thromb$score_multiomic_z, probs = c(1/3, 2/3), na.rm = TRUE)

dat_joint <- dat_thromb |>
  mutate(
    score_tertile = case_when(
      score_multiomic_z <= q_score[1] ~ "Low score",
      score_multiomic_z >= q_score[2] ~ "High score",
      TRUE ~ "Middle score"
    )
  ) |>
  filter(score_tertile != "Middle score") |>
  mutate(
    chip_score_group = case_when(
      CHIP.all == 0 & score_tertile == "Low score"  ~ "CHIP- / Low score",
      CHIP.all == 0 & score_tertile == "High score" ~ "CHIP- / High score",
      CHIP.all == 1 & score_tertile == "Low score"  ~ "CHIP+ / Low score",
      CHIP.all == 1 & score_tertile == "High score" ~ "CHIP+ / High score",
      TRUE ~ NA_character_
    ),
    chip_score_group = factor(
      chip_score_group,
      levels = c(
        "CHIP- / Low score",
        "CHIP- / High score",
        "CHIP+ / Low score",
        "CHIP+ / High score"
      )
    )
  ) |>
  filter(!is.na(chip_score_group))

table(dat_joint$chip_score_group, dat_joint$thromb)


fit_joint_thromb <- glm(
  reformulate(c("chip_score_group", var.arr), response = "thromb"),
  data = dat_joint,
  family = binomial()
)

joint_df <- broom::tidy(fit_joint_thromb, conf.int = TRUE, exponentiate = TRUE) |>
  filter(grepl("^chip_score_group", term)) |>
  mutate(
    group = gsub("^chip_score_group", "", term),
    group = factor(
      group,
      levels = c(
        "CHIP- / High score",
        "CHIP+ / Low score",
        "CHIP+ / High score"
      )
    )
  )

joint_plot <- bind_rows(
  data.frame(
    term = "Reference",
    estimate = 1,
    conf.low = 1,
    conf.high = 1,
    p.value = NA_real_,
    group = factor(
      "CHIP- / Low score",
      levels = c(
        "CHIP- / Low score",
        "CHIP- / High score",
        "CHIP+ / Low score",
        "CHIP+ / High score"
      )
    )
  ),
  joint_df |>
    mutate(
      group = factor(
        as.character(group),
        levels = c(
          "CHIP- / Low score",
          "CHIP- / High score",
          "CHIP+ / Low score",
          "CHIP+ / High score"
        )
      )
    )
)

joint_plot

p5F <- ggplot(joint_plot, aes(x = estimate, y = group)) +
  geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.18, linewidth = 0.7) +
  geom_point(size = 2.8) +
  scale_x_log10() +
  theme_classic(base_size = 13) +
  labs(
    x = "Adjusted OR for thromb",
    y = NULL
  )

p5F
```
### driver + score
```{r}
driver_vars <- c("DNMT3A.all", "TET2.all", "PPM1D.all", "ASXL1.all")

get_driver_thromb <- function(driver_var, dat, var.arr, add_score = FALSE) {
  
  terms_use <- c(driver_var, var.arr)
  if (add_score) {
    terms_use <- c(driver_var, "score_multiomic_z", var.arr)
  }
  
  fit <- glm(
    reformulate(terms_use, response = "thromb"),
    data = dat,
    family = binomial()
  )
  
  broom::tidy(fit, conf.int = TRUE, exponentiate = TRUE) |>
    filter(term == driver_var) |>
    mutate(
      driver = gsub("\\.all", "", driver_var),
      model = ifelse(add_score, "Adjusted for score", "Base model"),
      n = nobs(fit)
    )
}

driver_thromb_df <- bind_rows(
  lapply(driver_vars, get_driver_thromb, dat = dat_thromb, var.arr = var.arr, add_score = FALSE),
  lapply(driver_vars, get_driver_thromb, dat = dat_thromb, var.arr = var.arr, add_score = TRUE)
) |>
  mutate(
    driver = factor(driver, levels = c("DNMT3A", "TET2", "PPM1D", "ASXL1")),
    model = factor(model, levels = c("Base model", "Adjusted for score"))
  )

driver_thromb_df

p5G <- ggplot(driver_thromb_df, aes(x = estimate, y = driver, color = model)) +
  geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.5) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    position = position_dodge(width = 0.55),
    height = 0.18,
    linewidth = 0.7
  ) +
  geom_point(
    position = position_dodge(width = 0.55),
    size = 2.6
  ) +
  scale_x_log10() +
  theme_classic(base_size = 13) +
  labs(
    x = "Adjusted OR for thromb",
    y = NULL,
    color = NULL
  )

p5G
```


### score quantile 2

```{r}
fit_q <- glm(
  reformulate(c("score_q", var.arr), response = "thromb"),
  data = dat_thromb,
  family = binomial()
)

newdat <- dat_thromb |>
  summarise(across(all_of(var.arr), ~ {
    if (is.numeric(.x)) mean(.x, na.rm = TRUE) else levels(as.factor(.x))[1]
  }))

newdat <- newdat[rep(1, 4), ]
newdat$score_q <- factor(c("Q1", "Q2", "Q3", "Q4"), levels = c("Q1", "Q2", "Q3", "Q4"))

pred <- predict(fit_q, newdata = newdat, type = "link", se.fit = TRUE)

pred_df <- data.frame(
  score_q = newdat$score_q,
  fit = pred$fit,
  se = pred$se.fit
) |>
  mutate(
    prob = plogis(fit),
    lower = plogis(fit - 1.96 * se),
    upper = plogis(fit + 1.96 * se)
  )

p_f5_3 <- ggplot(pred_df, aes(x = score_q, y = prob * 100, group = 1)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.8) +
  geom_errorbar(aes(ymin = lower * 100, ymax = upper * 100), width = 0.12) +
  theme_classic(base_size = 13) +
  labs(
    x = "Integrated multi-omic score quartile",
    y = "Predicted thrombosis probability (%)"
  ) + 
  annotate(
    "text",
    x = 3.4,
    y = max(pred_df$upper * 100, na.rm = TRUE),
    label = paste0("P trend = ", trend_p$p.value),
    hjust = 0,
    size = 4
  )
p_f5_3
```

```{r}
library(dplyr)
library(purrr)
library(boot)

get_marginal_prob <- function(q_level, fit, dat) {
  newdat <- dat
  newdat$score_q <- factor(q_level,levels = c("Q1", "Q2", "Q3", "Q4"))
  pred_prob <- predict(fit,newdata = newdat,type = "response")
  tibble(score_q = q_level,prob = mean(pred_prob, na.rm = TRUE))
}

pred_marginal <- map_dfr(c("Q1", "Q2", "Q3", "Q4"),get_marginal_prob,fit = fit_q,dat = dat_thromb)

pred_marginal

set.seed(123)

boot_marginal <- function(dat, indices, var.arr) {
  d <- dat[indices, , drop = FALSE]
  fit <- glm(reformulate(c("score_q", var.arr), response = "thromb"),data = d,family = binomial())
  out <- map_dfr(c("Q1", "Q2", "Q3", "Q4"), function(q_level) {
    newdat <- d
    newdat$score_q <- factor(q_level,levels = c("Q1", "Q2", "Q3", "Q4"))
    tibble(score_q = q_level,prob = mean(predict(fit, newdata = newdat, type = "response"), na.rm = TRUE))
  })
  out$prob
}

boot_res <- boot(data = dat_thromb,statistic = function(data, indices) {boot_marginal(data, indices, var.arr)},R = 1000)

pred_marginal_ci <- pred_marginal |> mutate(
    lower = apply(boot_res$t, 2, quantile, probs = 0.025, na.rm = TRUE),
    upper = apply(boot_res$t, 2, quantile, probs = 0.975, na.rm = TRUE)
  )

pred_marginal_ci

p_f5_3_marginal <- ggplot(pred_marginal_ci,aes(x = score_q, y = prob * 100, group = 1)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2.8) +
  geom_errorbar(aes(ymin = lower * 100, ymax = upper * 100),width = 0.12) +
  theme_classic(base_size = 13) +
  labs(x = "Multi-layer score quartile",y = "Adjusted thrombosis risk (%)") +
  annotate("text",x = 3.25,y = max(pred_marginal_ci$upper * 100, na.rm = TRUE),
           label = "P for trend = 2.43e-6",hjust = 0,size = 4
  )

pdf("chip_f5c.pdf",width = 5, height =5)
p_f5_3_marginal
dev.off()
```

```{r}
dat_thromb$vaf_group <- factor(dat_thromb$vaf_group, levels = rev(c("CHIP-","Small_clone","Large_clone")))
p_f5_4 <- ggplot(dat_thromb, aes(x = vaf_group, y = score_multiomic_z)) +
  geom_violin(fill = "grey90", color = NA, width = 0.9) +
  geom_boxplot(width = 0.18, outlier.shape = NA, fill = "white") +
  theme_classic(base_size = 13) +
  labs(x = NULL,y = "Integrated multi-omic score") +
  stat_compare_means(method = "wilcox.test",paired = FALSE, label = "p.signif",
                     comparisons=list(c("Small_clone", "CHIP-"),c("Large_clone", "CHIP-")))
p_f5_4


ggplot(dat_thromb, aes(x = driver_group, y = score_multiomic_z)) +
  geom_violin(fill = "grey90", color = NA, width = 0.9) +
  geom_boxplot(width = 0.18, outlier.shape = NA, fill = "white") +
  theme_classic(base_size = 13) +
  labs(
    x = NULL,
    y = "Integrated multi-omic score"
  )
```

```{r}
library(ggalluvial)

flow_df <- dat_joint |>
  count(chip_score_group, thromb)

ggplot(flow_df,
       aes(axis1 = chip_score_group, axis2 = factor(thromb), y = n)) +
  geom_alluvium(aes(fill = chip_score_group), width = 1/12, alpha = 0.7) +
  geom_stratum(width = 1/12, fill = "grey90", color = "grey40") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_void()
```
```{r}
library(dplyr)
library(ggplot2)
library(ggalluvial)
library(scales)

flow_df_prop <- dat_joint |>
  count(chip_score_group, thromb) |>
  group_by(chip_score_group) |>
  mutate(
    total = sum(n),
    prop = n / total,
    label_prop = paste0(round(prop * 100, 1), "%")
  ) |>
  ungroup() |>
  mutate(
    thromb = factor(
      thromb,
      levels = c(0, 1),
      labels = c("No thrombosis", "Thrombosis")
    ),
    chip_score_group = factor(
      chip_score_group,
      levels = c(
        "CHIP- / Low score",
        "CHIP- / High score",
        "CHIP+ / Low score",
        "CHIP+ / High score"
      )
    )
  )

p_f5_5 <- ggplot(
  flow_df_prop,
  aes(
    axis1 = chip_score_group,
    axis2 = thromb,
    y = prop
  )
) +
  geom_alluvium(
    aes(fill = chip_score_group),
    width = 1/12,
    alpha = 0.75
  ) +
  geom_stratum(
    width = 1/12,
    fill = "grey90",
    color = "grey40"
  ) +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(stratum)),
    size = 4
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    y = "Within-group proportion",
    fill = NULL
  ) +
  theme_void() +
  theme(
    legend.position = "right"
  )

p_f5_5

ggplot(flow_df_prop, aes(x = chip_score_group, y = prop, fill = thromb)) +
  geom_col(width = 0.7, color = "white") +
  geom_text(
    aes(label = label_prop),
    position = position_stack(vjust = 0.5),
    size = 4
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = NULL,
    y = "Within-group proportion",
    fill = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1)
  )

event_rate_joint <- dat_joint |>
  group_by(chip_score_group) |>
  summarise(
    n = n(),
    events = sum(thromb == 1, na.rm = TRUE),
    event_rate = events / n,
    se = sqrt(event_rate * (1 - event_rate) / n),
    lower = event_rate - 1.96 * se,
    upper = event_rate + 1.96 * se,
    .groups = "drop"
  ) |>
  mutate(
    chip_score_group = factor(
      chip_score_group,
      levels = c(
        "CHIP- / Low score",
        "CHIP- / High score",
        "CHIP+ / Low score",
        "CHIP+ / High score"
      )
    )
  )

p_f5_6 <- ggplot(event_rate_joint, aes(x = chip_score_group, y = event_rate * 100)) +
  geom_col(width = 0.65, fill = "grey75", color = "black") +
  geom_errorbar(
    aes(ymin = lower * 100, ymax = upper * 100),
    width = 0.15
  ) +
  theme_classic(base_size = 13) +
  labs(
    x = NULL,
    y = "Thrombotic event rate (%)"
  ) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1)
  )
p_f5_6
```


```{r}
pdf("chip_f5b.pdf",width = 6, height = 6)
p_f5_1
p_f5_2
p_f5_3
p_f5_4
p_f5_5
p_f5_6
dev.off()

```


### score + medicine
```{r}
dat_med <- combind_dat |>
  mutate(
    CHIP_status = ifelse(CHIP.all == 1, "CHIP+", "CHIP-"),
    score_group = ifelse(
      score_multiomic_z >= median(score_multiomic_z, na.rm = TRUE),
      "High score", "Low score"
    ),
    chip_score_group = paste(CHIP_status, score_group, sep = " / "),
    chip_score_group = factor(
      chip_score_group,
      levels = c(
        "CHIP- / Low score",
        "CHIP+ / Low score",
        "CHIP- / High score",
        "CHIP+ / High score"
      )
    )
  )
```


```{r}
meds <- c("statin", "antidiabetic", "antithrombotic")

med_prop_df <- dat_med |>
  select(chip_score_group, all_of(meds)) |>
  tidyr::pivot_longer(
    cols = all_of(meds),
    names_to = "Medication",
    values_to = "Use"
  ) |>
  group_by(chip_score_group, Medication) |>
  summarise(
    n = n(),
    users = sum(Use == 1, na.rm = TRUE),
    prop = users / n,
    .groups = "drop"
  )
```

```{r}
p_5_b2_1 <- med_prop_df |>
  mutate(
    Medication = recode(
      Medication,
      statin = "Statin",
      antidiabetic = "Antidiabetic",
      antithrombotic = "Antithrombotic"
    )
  ) |>
  ggplot(aes(x = chip_score_group, y = prop, fill = Medication)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_test() +
  labs(
    x = NULL,
    y = "Medication use (%)",
    fill = NULL
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p_5_b2_1
```
```{r}
covars_base <- c(
  "Age", "Sex", "Smoking", "BMI",
  paste0("pc", 1:10)
)

med_covars <- c("statin", "antidiabetic", "antithrombotic")

run_group_model_compare <- function(outcome, data = dat_med) {
  
  fml_base <- as.formula(
    paste0(
      outcome, " ~ chip_score_group + ",
      paste(covars_base, collapse = " + ")
    )
  )
  
  fml_med <- as.formula(
    paste0(
      outcome, " ~ chip_score_group + ",
      paste(c(covars_base, med_covars), collapse = " + ")
    )
  )
  
  fit_base <- glm(fml_base, data = data, family = binomial())
  fit_med  <- glm(fml_med,  data = data, family = binomial())
  
  bind_rows(
    broom::tidy(fit_base, conf.int = TRUE, exponentiate = TRUE) |>
      filter(grepl("^chip_score_group", term)) |>
      mutate(model = "Base model"),
    
    broom::tidy(fit_med, conf.int = TRUE, exponentiate = TRUE) |>
      filter(grepl("^chip_score_group", term)) |>
      mutate(model = "Medication-adjusted model")
  ) |>
    mutate(outcome = outcome)
}

group_compare_df <- purrr::map_dfr(
  c("thromb", "VTE", "ATE"),
  run_group_model_compare
)
```

```{r}
group_compare_df <- group_compare_df |>
  mutate(
    outcome = recode(
      outcome,
      thromb = "Thrombosis",
      VTE = "VTE",
      ATE = "ATE"
    ),
    group = gsub("^chip_score_group", "", term),
    group = factor(
      group,
      levels = c(
        "CHIP+ / Low score",
        "CHIP- / High score",
        "CHIP+ / High score"
      )
    ),
    model = factor(model, levels = c("Base model", "Medication-adjusted model"))
  )
```

```{r}
group_compare_df |>
  ggplot(aes(x = estimate, y = group, color = model)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point(position = position_dodge(width = 0.5), size = 2.5) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    position = position_dodge(width = 0.5),
    height = 0.2
  ) +
  facet_wrap(~ outcome, nrow = 1) +
  scale_x_log10() +
  theme_test() +
  labs(
    x = "Adjusted OR versus CHIP- / Low score",
    y = NULL,
    color = NULL
  )

```
```{r}
attenuation_group_df <- group_compare_df |>
  select(outcome, group, model, estimate) |>
  tidyr::pivot_wider(names_from = model, values_from = estimate) |>
  mutate(
    attenuation_percent =
      (`Base model` - `Medication-adjusted model`) /
      (`Base model` - 1) * 100
  )

attenuation_group_df
```
```{r}
filter(attenuation_group_df, group == "CHIP+ / High score")
```
```{r}
library(ggeffects)

fit_statin <- glm(
  thromb ~ chip_score_group * statin +
    Age + Sex + Smoking + BMI +
    pc1 + pc2 + pc3 + pc4 + pc5 + pc6 + pc7 + pc8 + pc9 + pc10 +
    antidiabetic + antithrombotic,
  data = dat_med,
  family = binomial()
)

pred_statin <- ggpredict(fit_statin, terms = c("chip_score_group", "statin"))
```

```{r}
plot_med_pred <- function(med, outcome = "thromb", data = dat_med) {
  
  other_meds <- setdiff(med_covars, med)
  
  fml <- as.formula(
    paste0(
      outcome, " ~ chip_score_group * ", med, " + ",
      paste(c(covars_base, other_meds), collapse = " + ")
    )
  )
  
  fit <- glm(fml, data = data, family = binomial())
  
  ggpredict(fit, terms = c("chip_score_group", med)) |>
    as.data.frame() |>
    mutate(
      Medication = med,
      Med_status = ifelse(group == "1", "User", "Non-user")
    )
}

pred_med_group_df <- purrr::map_dfr(meds, plot_med_pred)
```
```{r}
p_5_b2_2 <- pred_med_group_df |>
  mutate(
    Medication = recode(
      Medication,
      statin = "Statin",
      antidiabetic = "Antidiabetic",
      antithrombotic = "Antithrombotic"
    )
  ) |>
  ggplot(aes(x = x, y = predicted, color = Med_status, group = Med_status)) +
  geom_point(position = position_dodge(width = 0.35), size = 2.5) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    position = position_dodge(width = 0.35),
    width = 0.15
  ) +
  geom_line(position = position_dodge(width = 0.35), linewidth = 0.7) +
  facet_wrap(~ Medication) +
  theme_test() +
  labs(
    x = NULL,
    y = "Adjusted predicted probability of thrombosis",
    color = NULL
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p_5_b2_2
```



```{r}
library(dplyr)

dat_med2 <- combind_dat |>
  mutate(
    CHIP_status = ifelse(CHIP.all == 1, "CHIP+", "CHIP-"),
    score_group = ifelse(
      score_multiomic_z >= median(score_multiomic_z, na.rm = TRUE),
      "High score", "Low score"
    ),
    vascular_med = ifelse(statin == 1 | antithrombotic == 1, 1, 0),
    med_group = ifelse(vascular_med == 1, "Medication", "No medication"),
    chip_score_med_group = paste(CHIP_status, score_group, med_group, sep = " / ")
  )
```

```{r}
event_rate_df <- dat_med2 |>
  group_by(chip_score_med_group) |>
  summarise(
    n = n(),
    events = sum(thromb == 1, na.rm = TRUE),
    event_rate = events / n,
    .groups = "drop"
  )
```

```{r}
event_rate_df |>
  ggplot(aes(x = reorder(chip_score_med_group, event_rate), y = event_rate)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = paste0(round(event_rate * 100, 1), "%")),
    hjust = -0.2,
    size = 3.5
  ) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(), expand = expansion(mult = c(0, 0.12))) +
  theme_test() +
  labs(
    x = NULL,
    y = "Thrombotic event rate"
  )
```
#### cr
```{r}
cr_time_df <- competing_risk.df |> select(eid, time, status, end_date)

dat_fig5_cr <- combind_dat |>
  left_join(cr_time_df, by = "eid")
dat_fig5_cr$status <- dat_fig5_cr$status.y

dat_fig5_cr |>
  summarise(
    n = n(),
    missing_time = sum(is.na(time)),
    missing_status = sum(is.na(status))
  )

dat_fig5_cr <- dat_fig5_cr |>
  mutate(
    time_year = time / 365.25,
    CHIP_status = ifelse(CHIP.all == 1, "CHIP+", "CHIP-"),
    score_group = ifelse(
      score_multiomic_z >= median(score_multiomic_z, na.rm = TRUE),
      "High score", "Low score"
    ),
    medication = ifelse(
      statin == 1 | antidiabetic == 1 | antithrombotic == 1,
      "Medication", "No medication"
    ),
    chip_score_group = paste(CHIP_status, score_group, sep = " / "),
    chip_score_med_group = paste(CHIP_status, score_group, medication, sep = " / ")
  )
```

```{r}
library(cmprsk)

dat_fig5_cr <- dat_fig5_cr |>
  mutate(
    chip_score_group = factor(
      chip_score_group,
      levels = c(
        "CHIP- / Low score",
        "CHIP+ / Low score",
        "CHIP- / High score",
        "CHIP+ / High score"
      )
    )
  )

ci_score <- cuminc(
  ftime = dat_fig5_cr$time_year,
  fstatus = dat_fig5_cr$status,
  group = dat_fig5_cr$chip_score_group,
  cencode = 0
)

plot(ci_score)
```


```{r}
cuminc_to_df_event1 <- function(ci_obj) {
  names(ci_obj) |>
    keep(~ !.x %in% "Tests") |>
    keep(~ grepl(" 1$", .x)) |>
    map_dfr(function(nm) {
      tibble(
        group_event = nm,
        time = ci_obj[[nm]]$time,
        estimate = ci_obj[[nm]]$est,
        variance = ci_obj[[nm]]$var
      )
    }) |>
    mutate(
      group = sub(" 1$", "", group_event),
      se = sqrt(variance),
      conf.low = pmax(0, estimate - 1.96 * se),
      conf.high = pmin(1, estimate + 1.96 * se)
    )
}

```
```{r}
ci_score_df <- cuminc_to_df_event1(ci_score)

p_5_b2_3 <- ci_score_df |>
  ggplot(aes(x = time, y = estimate, color = group)) +
  geom_step(linewidth = 1.1) +
  theme_test() +
  labs(
    x = "Time in years",
    y = "Cumulative incidence of thrombotic events",
    color = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme(
    legend.position = "right"
  )

p_5_b2_3
```
```{r}
df <- dat_fig5_cr |>  filter(!is.na(eid), !is.na(time_year), !is.na(status), status %in% c(0, 1, 2), time_year >= 0)
ch <- "chip_score_group"
var.arr = c("Sex","Age","BMI","Smoking","pc1","pc2","pc3","pc4","pc5","pc6","pc7","pc8","pc9","pc10","statin","antidiabetic","antithrombotic")
  fg <- finegray(reformulate(c(ch, var.arr,"eid"), response = "Surv(time/365.25, status)"), 
               data=df, etype = 1)
  fit_fg_crude <- coxph(reformulate(c(ch), response = "Surv(fgstart, fgstop, fgstatus)"),
                     data = fg, weights = fgwt, ties = "efron", robust = TRUE,cluster = eid)
  fit_fg_adj <- coxph(reformulate(c(ch,var.arr), response = "Surv(fgstart, fgstop, fgstatus)"),
                     data = fg, weights = fgwt, ties = "efron", robust = TRUE,cluster = eid)
  
  ggforest(fit_fg_adj, data = fg, main = "sHR (Fine-Gray Model)")
  ggforest(fit_fg_crude, data = fg, main = "sHR (Fine-Gray Model)")
  
  
  
```

```{r}
extract_coxph_factor <- function(fit, var_prefix, label){
  s <- summary(fit)$coefficients
  
  rn <- rownames(s)
  idx <- grepl(paste0("^", var_prefix), rn)
  
  if (!any(idx)) {
    stop("No coefficient found for variable prefix: ", var_prefix,
         "\nAvailable coefficients are:\n",
         paste(rn, collapse = "\n"))
  }
  
  data.frame(
    Model = label,
    term = rn[idx],
    HR = s[idx, "exp(coef)"],
    lower = exp(s[idx, "coef"] - 1.96 * s[idx, "robust se"]),
    upper = exp(s[idx, "coef"] + 1.96 * s[idx, "robust se"]),
    p = s[idx, "Pr(>|z|)"],
    row.names = NULL
  )
}

  ch <- "chip_score_group"

df_plot <- bind_rows(
  extract_coxph_factor(fit_fg_crude, ch, "Unadjusted"),
  extract_coxph_factor(fit_fg_adj, ch, "Adjusted")
)


df_plot <- df_plot |>
  dplyr::mutate(
    Group = gsub(paste0("^", ch), "", term),
    Group = factor(
      Group,
      levels = c(
        "CHIP+ / Low score",
        "CHIP- / High score",
        "CHIP+ / High score"
      )
    ),
    p_lab = dplyr::case_when(
      p < 0.0001 ~ "****",
      p < 0.001  ~ "***",
      p < 0.01   ~ "**",
      p < 0.05   ~ "*",
      TRUE ~ "ns"
    ),
    label = paste0(
      round(HR, 2), " [",
      round(lower, 2), "-",
      round(upper, 2), "] ",
      p_lab
    ),
    Model = factor(Model, levels = c("Unadjusted", "Adjusted"))
  )

ref_df <- expand.grid(
  Model = levels(df_plot$Model),
  Group = "CHIP- / Low score"
) |>
  mutate(
    HR = 1,
    lower = 1,
    upper = 1,
    p = NA_real_,
    p_lab = "Ref.",
    label = "Ref."
  )

df_plot <- bind_rows(df_plot, ref_df) |>
  mutate(
    Group = factor(
      Group,
      levels = c(
        "CHIP- / Low score",
        "CHIP+ / Low score",
        "CHIP- / High score",
        "CHIP+ / High score"
      )
    )
  )
  
  
p_5_b2_4 <- ggplot(df_plot, aes(y = Group, x = HR, color = Model)) + 
  geom_point(
    position = position_dodge(width = 0.5),
    size = 3
  ) +
  geom_errorbarh(
    aes(xmin = lower, xmax = upper),
    position = position_dodge(width = 0.5),
    height = 0.18
  ) +
  geom_vline(
    xintercept = 1,
    linetype = 2,
    color = "grey50"
  ) +
  scale_x_log10() +
  scale_color_manual(values = mycol3[1:2]) + 
  geom_text(
    aes(x = upper * 1.08, label = label),
    position = position_dodge(width = 0.5),
    hjust = 0,
    size = 3,
    show.legend = FALSE
  ) +
  theme_test() +
  ylab(NULL) + 
  xlab("Subdistribution hazard ratio (sHR)") +
  coord_cartesian(xlim = c(
    min(df_plot$lower, na.rm = TRUE) * 0.85,
    max(df_plot$upper, na.rm = TRUE) * 1.45
  ))
p_5_b2_4
```

```{r}
library(dplyr)
library(purrr)
library(tibble)
library(stringr)

# 你想提取的年份
years_to_extract <- 1:15

# 提取 cuminc 对象中 event = 1，也就是 thrombosis 的曲线
extract_cif_at_years <- function(ci_obj, years = c(5, 10, 15), event_code = "1") {
  
  curve_names <- names(ci_obj)
  curve_names <- curve_names[!curve_names %in% "Tests"]
  
  # 只保留 event_code = 1 的曲线
  curve_names <- curve_names[str_detect(curve_names, paste0(" ", event_code, "$"))]
  
  map_dfr(curve_names, function(nm) {
    
    x <- ci_obj[[nm]]
    
    group_name <- str_remove(nm, paste0(" ", event_code, "$"))
    
    map_dfr(years, function(yr) {
      
      idx <- which(x$time <= yr)
      
      if (length(idx) == 0) {
        est <- 0
        var <- 0
      } else {
        est <- x$est[max(idx)]
        var <- x$var[max(idx)]
      }
      
      se <- sqrt(var)
      
      tibble(
        `CHIP-score group` = group_name,
        Year = yr,
        `Cumulative incidence` = est,
        `Cumulative incidence (%)` = est * 100,
        `Standard error` = se,
        `95% CI lower (%)` = pmax(0, est - 1.96 * se) * 100,
        `95% CI upper (%)` = pmin(1, est + 1.96 * se) * 100
      )
    })
  })
}

cif_year_df <- extract_cif_at_years(
  ci_obj = ci_score,
  years = years_to_extract,
  event_code = "1"
)

cif_year_df
```
```{r}
pdf("chip_f5d.pdf", width = 5, height = 3.5, useDingbats = FALSE)
p_5_b2_1
p_5_b2_2
p_5_b2_3
p_5_b2_4
dev.off()

```


```{r}
write.table(flow_df_prop, "chip_suppl_table_mlthrom.tsv", sep = "\t", quote = FALSE,row.names = FALSE)
write.table(med_prop_df, "chip_suppl_table_mlmed1.tsv", sep = "\t", quote = FALSE,row.names = FALSE)
write.table(pred_med_group_df, "chip_suppl_table_mlmed2.tsv", sep = "\t", quote = FALSE,row.names = FALSE)

write.table(cif_year_df, "chip_suppl_table_mlcif.tsv", sep = "\t", quote = FALSE,row.names = FALSE)
write.table(df_plot, "chip_suppl_table_mlfg.tsv", sep = "\t", quote = FALSE,row.names = FALSE)
```


### cr other
```{r}
run_signature_fg <- function(dat, outcome_df, outcome_name,
                             score_var = "score_multiomic_z",
                             chip_var = "CHIP.all",
                             id_var = "eid") {
  
  library(dplyr)
  library(survival)
  library(cmprsk)
  library(purrr)
  library(tibble)
  library(ggplot2)
  
  group_levels <- c(
    "CHIP- / Low score",
    "CHIP+ / Low score",
    "CHIP- / High score",
    "CHIP+ / High score"
  )
  
  group_cols <- c(
    "CHIP- / Low score"  = "#9ec0d2",
    "CHIP+ / Low score"  = "#df8d8d",
    "CHIP- / High score" = "#236898",
    "CHIP+ / High score" = "#96245b"
  )
  
  var_arr <- c(
    "Sex", "Age", "BMI", "Smoking",
    paste0("pc", 1:10),
    "statin", "antidiabetic", "antithrombotic"
  )
  
  dat_fg <- dat |>
    left_join(
      outcome_df |>
        transmute(
          eid,
          outcome_time = time,
          outcome_status = status
        ),
      by = "eid"
    ) |>
    mutate(
      time_year = outcome_time / 365.25,
      status = outcome_status,
      CHIP_status = ifelse(.data[[chip_var]] == 1, "CHIP+", "CHIP-"),
      score_group = ifelse(
        .data[[score_var]] >= median(.data[[score_var]], na.rm = TRUE),
        "High score", "Low score"
      ),
      chip_score_group = paste(CHIP_status, score_group, sep = " / "),
      chip_score_group = factor(
        chip_score_group,
        levels = group_levels
      )
    ) |>
    filter(
      !is.na(eid),
      !is.na(time_year),
      !is.na(status),
      status %in% c(0, 1, 2),
      time_year >= 0,
      !is.na(chip_score_group)
    )
  
  ci_obj <- cuminc(
    ftime = dat_fg$time_year,
    fstatus = dat_fg$status,
    group = dat_fg$chip_score_group,
    cencode = 0
  )
  
  ci_df <- names(ci_obj) |>
    purrr::keep(~ !.x %in% "Tests") |>
    purrr::keep(~ grepl(" 1$", .x)) |>
    purrr::map_dfr(function(nm) {
      tibble(
        group_event = nm,
        time = ci_obj[[nm]]$time,
        estimate = ci_obj[[nm]]$est,
        variance = ci_obj[[nm]]$var
      )
    }) |>
    mutate(
      group = sub(" 1$", "", group_event),
      se = sqrt(variance),
      conf.low = pmax(0, estimate - 1.96 * se),
      conf.high = pmin(1, estimate + 1.96 * se),
      group = factor(group, levels = group_levels)
    )
  
  fg <- finegray(
    reformulate(
      c("chip_score_group", var_arr, "eid"),
      response = "Surv(time_year, status)"
    ),
    data = dat_fg,
    etype = 1
  )
  
  fit_crude <- coxph(
    Surv(fgstart, fgstop, fgstatus) ~ chip_score_group,
    data = fg,
    weights = fgwt,
    ties = "efron",
    robust = TRUE,
    cluster = eid
  )
  
  fit_adj <- coxph(
    reformulate(
      c("chip_score_group", var_arr),
      response = "Surv(fgstart, fgstop, fgstatus)"
    ),
    data = fg,
    weights = fgwt,
    ties = "efron",
    robust = TRUE,
    cluster = eid
  )
  
  extract_fg <- function(fit, label) {
    s <- summary(fit)
    coef_df <- as.data.frame(s$coefficients)
    ci_df2 <- as.data.frame(s$conf.int)
    
    rn <- rownames(coef_df)
    idx <- grepl("^chip_score_group", rn)
    
    tibble(
      Model = label,
      term = rn[idx],
      HR = ci_df2[idx, "exp(coef)"],
      lower = ci_df2[idx, "lower .95"],
      upper = ci_df2[idx, "upper .95"],
      p = coef_df[idx, "Pr(>|z|)"]
    )
  }
  
  df_plot <- bind_rows(
    extract_fg(fit_crude, "Unadjusted"),
    extract_fg(fit_adj, "Adjusted")
  ) |>
    mutate(
      Group = gsub("^chip_score_group", "", term),
      Group = factor(Group, levels = group_levels[-1]),
      p_lab = case_when(
        p < 0.0001 ~ "****",
        p < 0.001  ~ "***",
        p < 0.01   ~ "**",
        p < 0.05   ~ "*",
        TRUE ~ "ns"
      ),
      label = paste0(round(HR, 2), " ", p_lab),
      Model = factor(Model, levels = c("Unadjusted", "Adjusted"))
    )
  
  ref_df <- expand.grid(
    Model = levels(df_plot$Model),
    Group = "CHIP- / Low score"
  ) |>
    mutate(
      HR = 1,
      lower = 1,
      upper = 1,
      p = NA_real_,
      p_lab = "Ref.",
      label = "Ref."
    )
  
  df_plot <- bind_rows(df_plot, ref_df) |>
    mutate(
      Group = factor(Group, levels = group_levels),
      Model = factor(Model, levels = c("Unadjusted", "Adjusted"))
    )
  
  p_ci <- ggplot(ci_df, aes(x = time, y = estimate, color = group)) +
    geom_step(linewidth = 1.1) +
    theme_test() +
    labs(
      title = outcome_name,
      x = "Time in years",
      y = paste0("Cumulative incidence of ", outcome_name),
      color = NULL
    ) +
    scale_color_manual(values = group_cols, drop = FALSE) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    theme(
      legend.position = "right"
    )
  
  p_fg <- ggplot(
    df_plot,
    aes(y = Group, x = HR, color = Group, shape = Model)
  ) +
    geom_point(
      position = position_dodge(width = 0.55),
      size = 3
    ) +
    geom_errorbarh(
      aes(xmin = lower, xmax = upper),
      position = position_dodge(width = 0.55),
      height = 0.18
    ) +
    geom_vline(
      xintercept = 1,
      linetype = 2,
      color = "grey50"
    ) +
    scale_x_log10() +
    scale_color_manual(values = group_cols, drop = FALSE) +
    scale_shape_manual(
      values = c("Unadjusted" = 16, "Adjusted" = 17)
    ) +
    geom_text(
      aes(x = upper * 1.08, label = label),
      position = position_dodge(width = 0.55),
      hjust = 0,
      size = 3,
      show.legend = FALSE
    ) +
    theme_test() +
    ylab(NULL) +
    xlab("Subdistribution hazard ratio (sHR)") +
    guides(
      color = "none",
      shape = guide_legend(title = NULL)
    ) +
    coord_cartesian(
      xlim = c(
        min(df_plot$lower, na.rm = TRUE) * 0.85,
        max(df_plot$upper, na.rm = TRUE) * 1.45
      )
    )
  
  list(
    data = dat_fg,
    cuminc = ci_obj,
    fit_crude = fit_crude,
    fit_adj = fit_adj,
    ci_plot_df = ci_df,
    fg_plot_df = df_plot,
    p_ci = p_ci,
    p_fg = p_fg
  )
}

make_cr_df <- function(icd10_df,
                       pheno_df,
                       icd_pattern,
                       lag_days = 180,
                       censor_date = as.Date("2024-12-31")) {
  
  event_df <- icd10_df |>
    dplyr::filter(grepl(icd_pattern, ICD10)) |>
    dplyr::mutate(date = as.Date(date)) |>
    dplyr::group_by(eid) |>
    dplyr::slice_min(order_by = date, n = 1, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::select(eid, ICD10, Diagnosis, event_date = date)
  
  cr_df <- pheno_df |>
    dplyr::left_join(event_df, by = "eid") |>
    dplyr::mutate(
      Sex = as.integer(Sex == "Male"),
      Smoking = as.integer(Smoking == "Yes"),
      assess_date = as.Date(Date),
      Date_Death = as.Date(Date_Death),
      
      prevalent_or_lag_event = !is.na(event_date) & 
        event_date <= assess_date + lag_days
    ) |>
    dplyr::filter(!prevalent_or_lag_event) |>
    dplyr::mutate(
      end_date = pmin(event_date, censor_date, Date_Death, na.rm = TRUE),
      time = as.numeric(end_date - assess_date),
      status_num = dplyr::case_when(
        !is.na(event_date) & event_date == end_date ~ 1L,
        !is.na(Date_Death) & Date_Death == end_date ~ 2L,
        TRUE ~ 0L
      ),
      status = factor(status_num, levels = c(0, 1, 2))
    ) |>
    dplyr::filter(!is.na(time), time >= 0)
  
  return(cr_df)
}
```

```{r}
icd_patterns <- list(
  COPD = "^J4[1-4]",                         # J41-J44
  Chronic_liver_disease = "^K7[0-7]",        # K70-K77
  Chronic_kidney_disease = "^N18",           # N18
  Myeloid_leukemia = "^C9[2-4]",             # C92-C94
  Heart_failure = "^I50"                     # I50
)

cr_list <- purrr::imap(
  icd_patterns,
  ~ make_cr_df(
    icd10_df = ukb.icd10.df,
    pheno_df = chip_mat.phen.df,
    icd_pattern = .x,
    lag_days = 180,
    censor_date = as.Date("2024-12-31")
  )
)

purrr::map(cr_list, ~ .x |> dplyr::count(status_num, CHIP.all))


```

```{r}
cr_plot.ls <- lapply(names(cr_list),function(x){
  res <- run_signature_fg(
  dat = combind_dat,
  outcome_df = cr_list[[x]],
  outcome_name = x
)
  res$p_ci | res$p_fg
})
cr_plot.ls


#pdf("chip_f5e.pdf", width = 10, height = 3.5, useDingbats = FALSE)
#cr_plot.ls
#dev.off()
```
```{r}
res_thromb <- run_signature_fg(
  dat = combind_dat,
  outcome_df =competing_risk.df,
  outcome_name = "thromb"
)
pdf("chip_f5f.pdf", width = 10, height = 3.5, useDingbats = FALSE)
res_thromb$p_ci | res_thromb$p_fg
dev.off()
```

## Appendix
```{r}
#write.csv(chip_mat.phen.df,"chip_mat.phen.csv", quote = FALSE, row.names = FALSE)
#write.table(ukb.gwas_pheno.df,"GWAS_data/ukb.gwas_pheno.chip_thromb.tsv", quote = FALSE, sep = "\t", row.names = FALSE)
#write.table(olink.or, "ukb.olink.odd_ratio.tsv", quote = FALSE, sep = "\t", row.names = FALSE)
#write.table(nmr_meta.or, "ukb.nmr_meta.odd_ratio.tsv", quote = FALSE, sep = "\t", row.names = FALSE)
#write.table(ukb.icd10.df, "ukb.icd10.long.tsv", quote = FALSE, sep = "\t", row.names = FALSE)

#olink.or <- read.csv("ukb.olink.odd_ratio.tsv", sep = "\t")
#chip_mat.phen.df <- read.csv("chip_mat.phen.csv")
#ukb.gwas_pheno.df <- read.csv("GWAS_data/ukb.gwas_pheno.chip_thromb.tsv", sep = "\t")

#save.image(file = "UKB_CHIP_Olink.Rdata")
load("UKB_CHIP_Olink.Rdata")
```

```{r}
limma_olink_volcano <- function(limma_df,pv = 0.05, fc = 1.01,top = 5,gene.list = "",title){
  #gene.list <- c("FLT3")
  #top <- 10
  #fc <- 1.01
  #pv <- 0.05
  #gg <- chip.deps
  gg <- limma_df 
  gg$Protein <- rownames(gg)
  gg <- gg |> left_join(olink_anno.df, by = "Protein")
  gg <- gg |> filter(! is.na(logFC), !is.na(adj.P.Val))
  gene.list <- c((gg |> filter(logFC > 0) |> top_n(n=-top, wt=P.Value))$Symbol,gene.list)
  gene.list <- c((gg |> filter(logFC < 0) |> top_n(n=-top, wt=P.Value))$Symbol,gene.list)
  index <- match(gene.list,gg$Symbol)
  index <- na.omit(index)
  gg$group = "no"
  gg[gg$logFC > log2(fc) & gg$P.Value < pv,]$group <- "up"
  gg[gg$logFC < -log2(fc) & gg$P.Value < pv,]$group <- "down"
  gg$color <- gg$group
  gg$color[index] <- "black"
  mycolour = c("grey", "#A81E2C","#08537C", "black")
  names(mycolour) = c("no", "up", "down", "black")
  gg$label <- ""
  gg$label[index] <- gg$Symbol[index]
  gg[gg$group == "no",]$label <- ""
  #p <- ggplot(gg, aes(x = logFC, y = -log10(adj.P.Val)))
  p <- ggplot(gg, aes(x = logFC, y = -log10(P.Value)))
  p <- p + geom_point_rast(aes(color = group,size = abs(t)), shape = 16, alpha = 0.6, show.legend = TRUE)
  p <- p + geom_point(color = ifelse(gg$label == "" ,NA,"black"), aes(size = abs(t)),shape = 1, show.legend = FALSE)
  p = p + scale_color_manual(values = mycolour)
  p = p + scale_fill_manual(values = mycolour)
  p = p + geom_hline(yintercept = -log10(pv), linetype = "dotted")
  p = p + geom_vline(xintercept = c(-log2(fc), log2(fc)), linetype = "dotted")
  p = p + ggrepel::geom_text_repel(aes(label=label,color=group),show.legend = FALSE,fontface = "bold", size = 2.5,box.padding = unit(0.8, "lines"),point.padding = unit(0.3, "lines"), segment.size = 0.3, max.overlaps = 10000)
  p = p + theme_test() + ggtitle(title)
  return(p)
}

limma_bloodcount_volcano <- function(limma_df,title,pv = 0.01, fc = 1, pt.size = 3){
  #fc <- 1
  #pv <- 0.01
  gg <- limma_df
  gg$Flag <- rownames(gg)
  index <- match(rownames(gg),data_dict_df.blood_count$ent_id)
  gg$Event <- data_dict_df.blood_count$ent_name_s[index]
  gg <- gg |> filter(! is.na(logFC), !is.na(adj.P.Val))
  gg$group = "no"
  gg[gg$logFC > log2(fc) & gg$adj.P.Val < pv,]$group <- "up"
  gg[gg$logFC < -log2(fc) & gg$adj.P.Val < pv,]$group <- "down"
  mycolour = c("grey", "#A81E2C","#08537C")
  names(mycolour) = c("no", "up", "down")
  gg$label <- ""
  gg$label <- gg$Event
  gg[gg$group == "no",]$label <- ""
  p <- ggplot(gg, aes(x = logFC, y = -log10(adj.P.Val)))
  p <- p + geom_point(aes(color = group), shape = 16, alpha = 0.6, size = pt.size, show.legend = FALSE)
  p <- p + geom_point(color = ifelse(gg$label == "" ,NA,"black"), shape = 1, size = pt.size, show.legend = FALSE)
  p = p + scale_color_manual(values = mycolour)
  p = p + scale_fill_manual(values = mycolour)
  p = p + geom_hline(yintercept = -log10(pv), linetype = "dotted")
  p = p + geom_vline(xintercept = c(-log2(fc), log2(fc)), linetype = "dotted")
  p = p + ggrepel::geom_text_repel(aes(label=label,color=group),show.legend = FALSE,fontface = "bold", size = 2.5,box.padding = unit(0.8, "lines"),point.padding = unit(0.3, "lines"), segment.size = 0.3, max.overlaps = 10000)
  p = p + theme_test() + ggtitle(title) + xlab("Standardized mean difference (SD units)")
  return(p)
}

limma_metabolic_volcano <- function(limma_df,pv = 0.05, fc = 1,top = 5,meta.list="",title){
  #fc <- 1.00
  #pv <- 0.05
  #top <- 5
  #meta.list <- ""
  #gg <- topTable(fit, coef="CHIPCH+",number = Inf)
  gg <- limma_df 
  gg$meta <- rownames(gg)
  gg <- gg |> filter(! is.na(logFC), !is.na(adj.P.Val))
  meta.list <- c((gg |> filter(logFC > 0) |> top_n(n=-top, wt=P.Value))$meta,meta.list)
  meta.list <- c((gg |> filter(logFC < 0) |> top_n(n=-top, wt=P.Value))$meta,meta.list)
  index <- match(meta.list,gg$meta)
  index <- na.omit(index)
  gg$group = "no"
  gg[gg$logFC > log2(fc) & gg$adj.P.Val < pv,]$group <- "up"
  gg[gg$logFC < -log2(fc) & gg$adj.P.Val < pv,]$group <- "down"
  gg$color <- gg$group
  gg$color[index] <- "black"
  mycolour = c("grey", "#A81E2C","#08537C", "black")
  names(mycolour) = c("no", "up", "down", "black")
  gg$label <- ""
  gg$label[index] <- gg$meta[index]
  gg[gg$group == "no",]$label <- ""
  p <- ggplot(gg, aes(x = logFC, y = -log10(adj.P.Val)))
  p <- p + geom_point_rast(aes(color = group), shape = 16, alpha = 0.6, show.legend = FALSE)
  p <- p + geom_point(color = ifelse(gg$label == "" ,NA,"black"), shape = 1, show.legend = FALSE)
  p = p + scale_color_manual(values = mycolour)
  p = p + scale_fill_manual(values = mycolour)
  p = p + geom_hline(yintercept = -log10(pv), linetype = "dotted")
  p = p + geom_vline(xintercept = c(-log2(fc), log2(fc)), linetype = "dotted")
  p = p + ggrepel::geom_text_repel(aes(label=label,color=group),show.legend = FALSE,fontface = "bold", size = 2.5,box.padding = unit(0.8, "lines"),point.padding = unit(0.3, "lines"), segment.size = 0.3, max.overlaps = 10000)
  p = p + theme_test() + ggtitle(title) + xlab("Standardized mean difference (SD units)")
  return(p)
}

```
